---
title: "Loops in R"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    highlight: tango
runtime: shiny_prerendered
tutorial:
  id: "com.rstudio.primers.loops"
  version: 1.0
---

```{r setup, include=FALSE, message = FALSE, warning=FALSE}
library(learnr)

checker <- function(label, user_code, check_code, envir_result, evaluate_result, ...) {
  list(message = check_code, correct = TRUE, location = "append")
}
tutorial_options(exercise.timelimit = 60, exercise.checker = checker)
knitr::opts_chunk$set(error = TRUE, out.width = "100%")
```

```{r, context="server-start", include=FALSE}
# Capture metrics only if running an official primer hosted by RStudio
library(curl)
library(later)
tryCatch(
  source("https://metrics.rstudioprimers.com/learnr/installMetrics", local=TRUE), 
  error = function(e){ 
    warning("An error occurred with the tracking code.")
    print(e)
  }
)
```

```{r, context = "render", results = 'asis', echo = FALSE}
tryCatch(
  source("https://metrics.rstudioprimers.com/learnr/installClient", local=TRUE)$value,
  error = function(e){ 
    print("Warning: An error occurred with the client code.")
  }
)
```

## repeat

###  

Many Data Science tasks require you to do the same thing over and over again. This is boring work---but not for your computer!

But how do you ask your computer to repeat a task? 

* The purrr package provides one way. I recommend that you check out the Iteration primer to learn how to use purrr and its map functions.

* This tutorial will show you another way. You will learn how to repeat tasks with R's loop functions. You know know enough to understand loops, and loops will expand your ability to write useful functions.

### Did you know?

Did you know that some languages, like C, C++ and python, use loops as an _all purpose_ programming tool. This approach will not work well in R.

R's loops are the right tool for a very specific job, but they lose out to other methods for other jobs. This is due to R's user-oriented design: most R functions implement their own pre-optimized loops behind the scenes in a lower level language. These built-in loops are much faster than any loop you could write at the command line. 

But don't let this drive you loopy! You'll learn when and when not to use loops in R if you read through to the end of _When to use loops_. But first, we have a more pressing topic:

What is a loop anyways?

### loops

R contains three types of loops

1. `repeat`
1. `while` 
1. `for`

The simplest of these are `repeat` loops.

### repeat 

`repeat` repeats an expression over and over again. To use `repeat`, type `repeat` without trailing parentheses. Then use braces to enclose one or more lines of code, e.g.

```{r eval = FALSE}
repeat {
  print("Hello")
}
```

```{r echo = FALSE}
cat("Hello\nHello\nHello\nHello\nHello\n...")
```

`repeat` will execute all of the code between the braces. Then it will execute it again. And again. And again...

###  

```{r repeat-quiz, echo = FALSE}
question("When will `repeat` stop executing the code?",
         answer("After five times.", message = "Repeat will repeat your code until the end of time (I shortened things in the example above). This isn't particularly useful, so you will almost always use repeat with..."),
         answer("Never, unless you or your computer intervene.", correct = TRUE, 
         message = "Repeat will repeat your code until the end of time. This isn't particularly useful, so you will almost always use repeat with..."))
```

### break

When an R loop encounters the `break` command, it exits the loop. This lets you schedule the end of a `repeat` loop.

* How many times will R repeat the code in the loop below? Make a prediction then Click Submit Answer to see if you are right.

```{r repeat, exercise = TRUE}
n <- 1
repeat {
  print(n)
  if (n == 5) break
  n <- n + 1
}
```

```{r repeat-solution}
n <- 1
repeat {
  print(n)
  if (n == 5) break
  n <- n + 1
}
```

```{r repeat-check}
"R repeats the print statement five times and then exits the loop. Notice how we created an n variable to use with the loop. The loop reaches the end because we increment n by one each time we run the code. What would happen if you left out n <- n + 1? Don't try! The loop would print 1 for a very long time."
```

###   

Here is a good question: 

What happens to the value of `n` when you run the loop? 

`n` is defined as `1` _outside_ of the loop. Is `n` still `1` when the loop is finished?

```{r results='hide'}
n <- 1
repeat {
  print(n)
  if (n == 5) break
  n <- n + 1
}
n
```

```{r n-setup}
n <- 1
repeat {
  if (n == 5) break
  n <- n + 1
}
```

* Click Submit Answer to find out.

```{r n, exercise = TRUE}
n
```

```{r n-solution}
n
```

```{r n-check}
"n = 5. Are you surprised? R's loops are not like functions, which safely execute code in a new environment. R's loops execute code in the environment that calls them, which means that loops will modify the objects in the environment. This is important to know. It can be very useful---but not if you forget that it happens."
```

### count

One way to prevent loops from altering objects in your environment is to put the loop into a function. For example, here is a function that counts from one to the number `x`. Do you see how it works?

```{r}
count <- function(x) {
  n <- 1
  repeat {
    print(n)
    if (n == x) break
    n <- n + 1
  }
}
```

* Can you write something similar? Write a function named `count_down` that counts from `x` to one. Be sure to arrange for the loop to end! Then click Submit Answer.

```{r countdown, exercise = TRUE}
count_down <- function(x) {

  
  
  
  
  
}
```

```{r countdown-hint-1}
"Begin by setting n = x. In which direction should you increment n to get to one?"
```

```{r countdown-hint-2}
"Increment n by subtracting one from it on each repetition. You will want to do this _after_ you print the value of n for that repetition."
```

```{r countdown-hint-3}
"Arrange to break the loop after you print n when n = 1."
```

```{r countdown-solution}
count_down <- function(x) {
  n <- x
  repeat {
    print(n)
    if (n == 1) break
    n <- n - 1
  }
}
```

```{r countdown-check}
"You know, this sounds a lot like a rocket launch. Let's try to make the loop return the words 'Blast off!' when it ends. If we can do that, we could make loops return _anything_ when they end."
```

### return

`break` is not the only way to end a loop in R. 

If your loop is in a function, you can end the loop with a `return()` statement. R will stop executing the function (and therefore the loop) and return the value supplied by `return()`. 

* Alter your function below to end the loop with `return("Blast off!")`. You will no longer need the `break` command. Then click Submit Answer.

```{r blastoff, exercise = TRUE}
count_down <- function(x) {
  n <- x
  repeat {
    print(n)
    if (n == 1) break
    n <- n - 1
  }
}
```

```{r blastoff-hint}
"Replace `break` with something else."
```

```{r blastoff-solution}
count_down <- function(x) {
  n <- x
  repeat {
    print(n)
    if (n == 1) return("Blast off!")
    n <- n - 1
  }
}
```

```{r blastoff-check}
"Return is a useful way to return a message or value from a loop within a function. And for satisfaction's sake..."
```

###  

```{r echo = FALSE}
count_down <- function(x) {
  n <- x
  repeat {
    print(n)
    if (n == 1) return("Blast off!")
    n <- n - 1
  }
}
```

```{r}
count_down(10)
```

Congratulations! 

A loop executes a piece of code multiple times, perhaps allowing changes to accumulate as it goes.

###  

You've learned the essence of loops, now it is time to learn the forms. In the next sections, you will learn how to refine `repeat` loops into two common types of loops: 
1. `while` loops, and
1. `for` loops

Each of these is a `repeat` loop adapted to a specific task.

###  

Along the way, you will work through a simple project: you will write and then refine a function that checks whether a number is prime. (This is a common computer science right of passage).

Let's begin the project now with a `repeat` loop. You'll need to know some things about prime numbers and `%%` to begin...

###   Prime numbers

In mathematics, a _prime_ number is a number that can be divided evenly only by itself and the number one. In other words, if you try to divide a prime number by any number less than itself, you get a remainder (_to keep things simple, let's not worry about zero or negative numbers_). 

For example, five is a prime number because you get a remainder when you divide it by two, three, and four:

```{r results ='hold'}
5 / 5
5 / 4
5 / 3
5 / 2
5 / 1
```

Interestingly, it is hard to prove that a number is prime...

###  

...unless you are a computer. Then you can divide the number by every number less than itself and show that the result has a remainder.

###  %%

To do this, you'll need to use the modulo function, `a %% b`, which we met in the control flow tutorial. Modulo is an arithmetic operator that returns the remainder of dividing `a` by `b`. For example, when you divide five by five, nothing is left over. When you divide five by four, one is left over. When you divide five by three, two is left over. And so on.

```{r results='hold'}
5 %% 5
5 %% 4
5 %% 3
5 %% 2
5 %% 1
```

Modulo doesn't return a decimal, it returns the number that remains once you subtract the largest multiple of `b` from `a` (i.e the number that you would get for the remainder if you did long division).

### is_prime

The code below uses a `repeat` loop to provide the beginnings of a function. Complete the loop:

1. Change `n` to begin at `2` instead of `1`
1. Add an `if` statement that ends the loop and returns `TRUE` if/when `n == x`.
1. Add a second `if` statement that that checks whether `x %% n == 0` and returns `FALSE` if so.
1. Then Click Submit Answer

```{r prime-repeat, exercise = TRUE}
is_prime <- function(x) {
  n <- 1
  repeat {
    print(n)
    
    
    n <- n + 1
  }
}
```

```{r prime-repeat-solution}
is_prime <- function(x) {
  n <- 2
  repeat {
    print(n)
    if (n == x) return(TRUE)
    if (x %% n == 0) return(FALSE)
    n <- n + 1
  }
}
```

```{r prime-repeat-check}
"The function works because a non-prime number will trigger the second if statement and return FALSE before the loop progresses to n = x. A prime number will reach n = x and return TRUE."
```

###  

* Use `is_prime` to check whether or not 89 is a prime number. Click Submit Answer to run your code.

```{r make-prime-repeat}
is_prime <- function(x) {
  n <- 2
  repeat {
    print(n)
    if (n == x) return(TRUE)
    if (x %% n == 0) return(FALSE)
    n <- n + 1
  }
}
```

```{r eighty-nine, exercise = TRUE, exercise.setup = "make-prime-repeat"}

```

```{r eighty-nine-solution}
is_prime(89)
```

```{r eighty-nine-check}
"89 *is* prime. Now that you have a working prime function, let's modify it to use a more streamlined loop."
```

## while

###  

`while` loops run while a logical condition is true. 

And, hey! So does our `repeat` loop! It runs _while_ n is less than x. How do you know? Because it stops and returns `TRUE` when n equals x.

```{r eval = FALSE}
is_prime <- function(x) {
  n <- 2
  repeat {
    print(n)
    if (n == x) return(TRUE)
    if (x %% n == 0) return(FALSE)
    n <- n + 1
  }
}
```

Let's replace `repeat` with `while`.

###  

`while` takes a logical test, like `if`, and a chunk of code, like `repeat`. `while` will repeat the code until the logical test returns `FALSE` (unless something in the code ends the loop first). 

```{r}
n <- 1
while (n <= 5) {
  print(n)
  n <- n + 1
}
```

You can read `while` as "while this condition is true, repeat that."

### Can you use while?

* Rewrite `is_prime` to use `while` instead of `repeat`. The new loop should run while n is less than x. 
* Have `is_prime` return `TRUE` in its last line, after the loop is run. 
    * Under which conditions, will the function return `TRUE`?
* Then click Submit Answer.

```{r prime-while, exercise = TRUE}
is_prime <- function(x) {
  n <- 2
  repeat {
    print(n)
    if (n == x) return(TRUE)
    if (x %% n == 0) return(FALSE)
    n <- n + 1
  }
}
```

```{r prime-while-hint}
"The new loop should run while n is less than x."
```

```{r prime-while-solution}
is_prime <- function(x) {
  n <- 2
  while (n < x) {
    print(n)
    if (x %% n == 0) return(FALSE)
    n <- n + 1
  }
  TRUE
}
```

```{r prime-while-check}
"`is_prime` will return TRUE if the loop comes to a natural end. That will only happen if `x` is a prime number."
```

###  

Congratulations! Your `while` loop works.

```{r echo = FALSE}
is_prime <- function(x) {
  n <- 2
  while (n < x) {
    print(n)
    if (x %% n == 0) return(FALSE)
    n <- n + 1
  }
  TRUE
}
```

```{r}
is_prime(7)
```

In fact, it does the same thing as your `repeat` loop. So why use `while`? 

###  

`while` loops can be easier to write than `repeat` loops; you do not need to insert your own `if` statement. 

`while` loops are also easier to read: `while` makes the code author's intentions obvious.

Now let's look at `for`. `for` loops provide the same advantages for a different situation. A for loop says "for each of these values do this." 

## for

###  

A for loop takes a defined set of values and repeats the loop once for each of the values. 

And, hey! So does our while loop! It runs once for each value in the set {2, 3, 4,  ...x}. How do you know? Because the loop begins with n = 2, increments n by one, and ends when n = x. 

```{r eval = FALSE}
is_prime <- function(x) {
  n <- 2
  while (n < x) {
    print(n)
    if (x %% n == 0) return(FALSE)
    n <- n + 1
  }
  TRUE
}
```

Let's replace `while` with `for`.


### for syntax

The syntax of `for` is similar to `while`, but instead of taking a logical test, `for` takes a three part statement:

1. An object that may, or may not, be used in the code chunk that follows `for`.
2. `in` (this never changes)
3. A vector of values to iterate over. On each run of the loop, `for` will assign a different value of the vector to the object named in step 1.


```{r}
for (n in c(1, 2, 3, 4, 5)) {
  print(n)
}
```

`for` automatically ends the loop once it has iterated over each value in the vector.

### for example

`for` loops are very versatile. For example, `for` makes it easy to increment over strange sets of values:

```{r}
for (i in c(1, 10, 200)) {
  print(i)
}
```

### for the win

You can also increment over "un-ordered" sets. You don't even need to stick to numbers:

```{r}
for (person in c("Betty", "Veronica", "Archie")) {
  greeting <- paste("Hello", person)
  print(greeting)
}
```

### seq

That's great! But how can you pass `for` the set {2, 3, 4,  ...x}? You'll need to do something like that to retool `is_prime`.

One possibility is to use `seq(2, x)`, which creates a sequence of integers from 2 to x.

```{r}
seq(1, 10)
```

`seq`, `seq_len` and `seq_along` form a complete family of helper functions for creating sequences in R.

### Can you use for?

* Convert `is_prime` to use a for loop that iterates over `n in seq(2, x - 1)`. 
    * Why wouldn't you want the loop to go all the way to n = x?
* Then click Submit Answer.

```{r for-1, exercise = TRUE}
is_prime <- function(x) {
  n <- 2
  while (n < x) {
    print(n)
    if (x %% n == 0) return(FALSE)
    n <- n + 1
  }
  TRUE
}
```

```{r for-1-hint}
"You no longer need to define n with an initial value, nor do you need to increment n."
```

```{r for-1-solution}
is_prime <- function(x) {
  for (n in seq(2, x - 1)) {
    print(n)
    if (x %% n == 0) return(FALSE)
  }
  TRUE
}
```

```{r for-1-check}
"You don't want the loop to continue to the case where n = x because x %% x always returns zero, even if x is prime, which means our function would always return FALSE. That wouldn't be good."
```

###  

Here's an odd question: what if you wanted to skip some value in a loop?

For example, what if  we wanted to skip n = 5 when we run our loop?

You can do that with the last loop helper provided by R, `next`.

### next

When R encounters `next` in a loop, it will move on to the next iteration of the loop without executing the rest of the loop. Here, when n = 5, `next` causes R to move on to the next iteration of the loop (where n will equal 6).

```{r eval = FALSE}
is_prime <- function(x) {
  for (n in seq(2, x - 1)) {
    if (n == 5) next
    print(n)
    if (x %% n == 0) return(FALSE)
  }
  TRUE
}
```

If five is the last value in the loop, `next` will cause R to exit the loop.

### Quiz

```{r eval = FALSE}
n <- 1
while (n < 2) {
  print(n)
  next
  n <- n + 1
}
```

```{r next-quiz, echo = FALSE}
question("Will this loop run forever?",
         answer("No"),
         answer("Yes", correct = TRUE, message = "n will never increment to two  because R will skip over the end of the loop each time it encounters next. Be thoughtful when you write loops!"),
         allow_retry = TRUE)
```

###  

Congratulations! You've learned how to use all of R's loop functions! To be a loop master, click on the button to learn when to use loops in R and when not to.

## When to use loops

###  

I mentioned earlier that loops should not appear as frequently in your R code as they would in your C, C++, or python code. Why is this? Because R is an extremely vectorized language.

### Vectorization

By "vectorized" I mean that most R functions are designed to take vectors as input and to return vectors as output. So if, for example, you'd like to take the square root of every value in a vector, you do not need to loop over the values in the vector...

```{r}
x <- c(1, 2, 3, 4, 5)
y <- double(length = 5)
for (i in seq_along(x)) {
  y[i] <- sqrt(i)
}
y
```

...you can just pass the vector to the square root function.

```{r}
sqrt(x)
```

And the best part about R's vectorization is that it is very fast!

###  

R's arithmetic operators are also vectorized. If you add two vectors, R will add each pair of elements in the vectors and return the results as a new vector.

```{r}
c(1, 2, 3, 4, 5) + c(1, 2, 3, 4, 5)
```

###  

And if you use R functions to build your functions, then your functions will inherit R's vectorization:

```{r}
round_square <- function(vec) {
  sqs <- sqrt(vec)
  round(sqs, digits = 2)
}

round_square(x)
```

###   

Vectorization reduces the need to use loops. You can think of each vectorized function as implementing a loop for you.

### recursion

Some problems cannot be solved with vectorized functions, but even for these you do not necessarily need to use a loop. 

For example, we could solve the prime problem with recursion. This is the strategy of having a function call itself. Until one of the recursively recalled functions returns an answer that is then passed up the call stack.

```{r}
is_prime <- function(x, n = 2) {
  if (n == x) return(TRUE)
  else if (x %% n == 0) return(FALSE)
  else is_prime(x, n = n + 1)
}
```

```{r}
is_prime(89)
is_prime(88)
```

### map

Or you could use a map function from the purrr package.

It is all up to you. The moral is that loops are not an all purpose tool in R. They are best reserved for problems that are not easy to solve with a vectorized function.

To learn more about when you definitely _should_ use a loop, check out Hadley Wickham's three suggestions in [Advanced R](http://adv-r.had.co.nz/Functionals.html#functionals-not) (be aware, it is advanced).
