---
title: "How to write a function"
tutorial:
  version: 0.01
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
---

```{r setup, include=FALSE, message = FALSE, warning=FALSE}
library(learnr)
library(grader)

tutorial_options(exercise.timelimit = 60, exercise.checker = grade_learnr)
knitr::opts_chunk$set(error = TRUE, out.width = "100%")
```

## Functions

### \t

Functions are the verbs of R. When something happens in R, it happens by running a function. For example,

* To take a log, you run a function named log: `log(2.71)`.
* To fit a linear model, you run a function named lm: `lm(mpg ~ wt, data = mtcars)`.
* To add two numbers, you run a function named "+": `1 + 2`. 

This last one is a shortcut for the function `'+'(1, 2)`. Everything that does something in R is a function, even if it doesn't seem like it.

### \t

So what should you do if you want to do something new in R, something that doesn't yet have a function?

### \t

Write a function?

No, silly. Write some code. But what should you do if you want to use that code _over and over again_, as if it were a part of R?

### \t

Write a function!

This tutorial will show you how. Take the quiz below to begin.

### Quiz

Suppose you need to run the following code on 20 different objects, `sqrt(sum(x^2))`.

```{r why-quiz, echo = FALSE}
question("Why would you not want to rewrite the code 20 times? (Check all that apply)",
     answer("Retyping code wastes time.", correct = TRUE),
     answer("Typing the code 20 times creates 20 chances to make a typo.", correct = TRUE),
     answer("If you need to change the code later, you'll need to change it in 20 different places.", correct = TRUE),
     allow_retry = TRUE, 
     type = "multiple", 
     incorrect = "Did you check ALL that applied?",
     correct = "Good job! You can avoid these outcomes by turning your code into a function. A good rule is to turn a piece of code into a function whenever you find yourself re-using the code more than three times.")
```


## Best practices

To write a function in R, follow these four steps:

1. Create a real R object (or set of objects) to use with your function
2. Write code that works with the real object(s)
3. Wrap the code in `function()`
4. Assign the name(s) of your real objects as argument names to the function

Consider these steps as the "best practice" for writing functions in R: they assure that your functions will work, which means that you will spend more time using your functions and less time debugging them. 

Let's use the steps to create your first function.

### Goal - Grading

To make this real, put yourself in the shoes of a teacher: 

You've given your students 10 homework assignments and announced that you will drop their lowest grade. The final grade will be the average grade of the remaining homeworks.

To make your life easier, you want to write an R function that will take a vector of 10 homework scores and return a final grade. 

Ready to begin?

### \t

Remember our steps:

1. Create a real R object (or set of objects) to use with your function

* Create an object named `x` that contains the vector `c(100, 100, 100, 100, 100, 100, 100, 100, 100, 90)`. Then click Submit Answer. `x` will be the grades of our test student. 

```{r obj, exercise = TRUE}

```

```{r obj-solution}
x <- c(100, 100, 100, 100, 100, 100, 100, 100, 100, 90)
```

```{r obj-check}
strict_check("Now we can use x to develop and test our code")
```

### \t

2. Write code that works with the real object(s)

Recall the grading scheme: the final grade is the average grade of the homework scores after you drop the lowest score. Since there are many ways to calculate this, let's try to be specific and end up on the same page.

* Use `sum()`, `min()`, `/`, `9` and parentheses to calculate the final grade for student `x`. Click Submit Answer to check that your code works.

```{r make-x, echo = FALSE}
x <- c(100, 100, 100, 100, 100, 100, 100, 100, 100, 90)
```

```{r norm, exercise = TRUE, setup = "make-x"}
x
```

```{r norm-hint}
"Your code should return 100."
```

```{r norm-solution}
(sum(x) - min(x)) / 9
```

```{r norm-check}
strict_check("This is the crucial step in writing a function. It is important to get your code working in real life with a real object before you try to abstract the code into a reusable function.")
```

### \t

3. Wrap the code in `function()`

You can use the `function()` function to wrap any piece of code into a function. To do this, place an opening bracket `{` after `function()`, then place your code, then place a closing bracket `}`, e.g.

```{r}
foo <- function() {
 a <- 1
 b <- 2
 a + b
}
```

You can place as many lines of code between the braces as you like. `function()` will wrap everything that appears between the braces into an R function. If you'd like to save the function, you'll need to assign it in the usual way to an R object that you can call later.

### \t

Once you've saved a function, you can inspect its contents by calling the name of the function. This is the function that R saved in the object named `foo`.

```{r}
foo
```

### \t

You can run the function by calling the name of the funciton followed by an open and closed parenthesis. The parentheses act as a trigger that causes R to run the code contained in the function.

```{r}
foo()
```

### \t

Let's save your code as a function.

* Save the code below as a function named `grade`. Then click Submit Answer.

```{r grade, exercise = TRUE}
(sum(x) - min(x)) / 9
```

```{r grade-hint}
"Use the function() {} function."
```

```{r grade-solution}
grade <- function() { 
  (sum(x) - min(x)) / 9
}
```

```{r grade-check}
strict_check("Now that you've saved grade as a function, you can call it whenever you like, but there is still one more thing to do...")
```

### \t

4. Assign the name(s) of your real objects as argument names to the function

At the moment, you `grade()` function is reusable but not _generalizable_. Each time you call `grade()` it computes the final grade of the vector `x` that contains `c(100, 100, 100, 100, 100, 100, 100, 100, 100, 90)`.

```{r echo = FALSE}
x <- c(100, 100, 100, 100, 100, 100, 100, 100, 100, 90)
grade <- function() { 
  (sum(x) - min(x)) / 9
}
```

```{r}
grade()
grade()
```

### \t

We'd like to use `grade()` with new vectors that have new values.

You can make any function in R generalizable by telling R that some objects in the function's code are _arguments_. An argument is an object that a user can assign a value to when he or she calls the function. The function will use the user's value for the object when it executes its code.

For example, we'd like to tell R that `x` in `grade()` is an argument. R shouldn't use a pre-defined value for `x`; it should let the user supply a new value for `x` each time he or she runs the function.

```{r eval = FALSE}
grade <- function() { 
  (sum(x) - min(x)) / 9
}
```

### \t

How do you tell R that an object is an argument? 

You list the name of the object in the parentheses that follow `function()` in the function definition. For example, you could make `a` and `b` arguments of my `foo` function.

```{r}
foo <- function(a, b) {
 a + b
}
```

If you make more than one argument, separate their names with a comma.

### \t

Now I can define a new value for `a` and `b` each time I call `foo`. 

```{r}
foo(a = 1, b = 1)
foo(a = 100, b = 200)
```

### \t

`foo` will always use my values of `a` and `b` to run the function---even if `a` and `b` are defined elsewhere. 

```{r}
a <- 5
b <- 6
foo(a = 2, b = 2)
```

### \t

Also, `foo()` will only use my values for that current run of the function. `foo()` won't overwrite the other definitions of `a` and `b`, and `foo` won't remember my values of `a` and `b` the next time I call `foo`.

```{r}
a
b
foo()
```

### \t

Interesting, huh? Now apply what you've learned to `grade()`.

* Change the code below to define `x` as an argument of `grade()`. Then click Submit Answer.

```{r x, exercise = TRUE}
grade <- function() { 
  (sum(x) - min(x)) / 9
}
```

```{r x-solution}
grade <- function(x) { 
  (sum(x) - min(x)) / 9
}
```

```{r x-check}
strict_check('If you use the "best practice" workflow, your argument names will always be the names of the real R objects that you create in Step 1. Nice and simple.')
```

### \t 

Good job! You can now have a finished `grade()` function that you can use to calculate the final grade of _any_ vector (I mean, student). Try it out.

* Calculate the final grade of the vector `c(100, 90, 90, 90, 90, 90, 90, 90, 90, 80)`. Then click Submit Answer.

```{r make-grade}
grade <- function(x) { 
  (sum(x) - min(x)) / 9
}
```

```{r function, exercise = TRUE, exercise.setup = "make-grade"}

```

```{r function-solution}
grade(x = c(100, 90, 90, 90, 90, 90, 90, 90, 90, 80))
```

```{r function-check}
strict_check("We'll learn more about arguments in the next tutorial. Let's recap the function writing workflow.")
```

### Quiz

```{r recap-quiz, echo = FALSE}
question("Which is not a step of the best practice workflow for writing functions?",
         allow_rety = TRUE, random_answer_order = TRUE,
         answer("Create a real R object (or set of objects) to use with your function"),
         answer("Write code that works with the real object(s)"),
         answer("Wrap the code in `function()`"),
         answer("Assign the name(s) of your real objects as argument names to the function"),
         answer("Predict how the function should be written and try it out.", correct = "This looks like a tempting way to save time, but more often than not it will lead to lengthy debugging sessions.")
)
```

### Recap

Use the four step workflow whenever you need to write a function:

1. Create a real R object (or set of objects) to use with your function
2. Write code that works with the real object(s)
3. Wrap the code in `function()`
4. Assign the name(s) of your real objects as argument names to the function

## Function look-alikes

### \t

If you've been using the tidyverse for a while, there's a good chance that you've already made things that resemble functions. 

Here are two examples that may be familiar. In each case, you can easily convert your code into a function.

### Purrr expressions

In the Iteration primer, you learned that purrr's map functions can take an "expression" to apply to each element of a vector. An _expression_ is a piece of code that is preceded by a `~` and includes `.x`s.

For example, this call to `map()` will apply the expression ``~.x^2` to each element of `vec`. On each iteration, `map()` will assign the _i_th element of vec to `.x` and run the expression.

```{r message=FALSE}
library(tidyverse)

vec <- c(1, 2, 3)
map(vec, ~.x^2)
```

### \t

In other words, purrr's map expressions work exactly like functions. To transform an expression to a function, remove the `~`, wrap the code in `function()` and list `.x` as an argument.

```{r}
sq <- function(.x) {
  .x^2
}

map(vec, sq)
```

### Functions and pipes

It is also easy to treat pipes as functions (since you can paste a pipe behind an input value). For example, this pipe would compute the L~2~ norm of any vector you place it behind.

```{r}
c(1, 1, 1) %>% 
  sq() %>% 
  sum() %>% 
  sqrt()

c(1, 2, 3, 4, 5) %>% 
  sq() %>% 
  sum() %>% 
  sqrt()
```

### \t 

It is a good idea to turn a pipe into a function when you find yourself using it often. You don't need to rewrite the pipe to do this. Pipes come with a handy shortcut for creating single argument functions.

To turn a pipe into a function, replace its initial object with a `.`, and then save the pipe to an object. R will treat the object as a function that passes its argument to the beginning of the pipe.

```{r}
l2_pipe <- . %>% 
  sq() %>% 
  sum() %>% 
  sqrt()

l2_pipe(c(1, 1, 1))
```

## Practice

Congratulations on finishing the tutorial. Before you go, complete the following exercises to cement your skills!

### Exercise 1

```{r e1, echo = FALSE}

question("What is the first step of writing a function?",
         answer("Choose a name, then type <- function(){", message = "You shouldn't type function() until you have code that works with a real test case."),
         answer("Create a real R object (or set of objects)", correct = TRUE, message = "You can then use the object to test your code."),
         answer("Assign argument names to your function"),
         answer("Write a help page that describes how your function will work."),
         allow_retry = TRUE)
```

### Exercise 2

Let's write a function that grades by counting a student's highest score twice and then taking an average of the 11 scores.

```{r e2, echo = FALSE}
question("Do you have a test vector that you can use?",
         answer("Test vector? We don't need no stinking test vector.", 
                message = "Yes you do."),
         answer("Yes, we can reuse the vector named x."),
         allow_retry = TRUE,
         random_answer_order = TRUE)
```

### Exercise 3

* Write a piece of code that uses `max()` to double count the highest score in `x` and then takes the average of the 11 results. Click Submit Answer to check that your code works.

```{r e3, exercise = TRUE, exercise.setup = "make-x"}
x
```

```{r e3-hint}
"Your code should retrun the answer 99.09091."
```

```{r e3-solution}
(sum(x) + max(x)) / 11
```

```{r e3-check}
strict_check("That student will be pleased!")
```

### Exercise 4

* Wrap your code in `function()` and save it to the name `grade2`.
* Then define the argument(s) for the function.
* Then click Submit Answer.

```{r e4, exercise = TRUE}
(sum(x) + max(x)) / 11
```

```{r e4-hint}
"When you use this workflow, the names of the argument(s) are always the names of the object(s) that you created in step 1."
```

```{r e4-solution}
grade2 <- function(x) {
  (sum(x) + max(x)) / 11
}
```

```{r e4-check}
strict_check("Your skills are almost complete.")
```

### Exercise 5

It's time to put it all together. 

* Use the code block below to develop a function named `l2` that calculates the Euclidean distance, or L~2~ norm ($\|x||_{2}$), of a vector of numbers named `x`. The L~2~ norm is the square root of the sum of the squared values in the vector, i.e.

$$\| x \|_{2} = \sqrt{\sum_{i = 1}^{n} x_{i}^{2}}$$

* When you are finished, reduce your code to just the definition of your function. Then click Submit Answer.

```{r e5, exercise = TRUE}

```

```{r e5-hint-1}
"You've already seen the code for the l2 function earlier in this tutorial (I do not mean the pipe example)."
```

```{r e5-hint-2}
"Consider using sqrt(), sum(), and ^2. Your code will work with vectors of any length."
```

```{r e5-solution}
l2 <- function(x) {
  sqrt(sum(x^2))
}
```

```{r e5-check}
strict_check()
```









