---
title: "Advanced Control Flow"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    highlight: tango
runtime: shiny_prerendered
tutorial:
  id: "com.rstudio.primers.advanced-control-flow"
  version: 1.0
---

```{r setup, include=FALSE, message = FALSE, warning=FALSE}
library(learnr)
library(lubridate)
library(tidyverse)

checker <- function(label, user_code, check_code, envir_result, evaluate_result, ...) {
  list(message = check_code, correct = TRUE, location = "append")
}
tutorial_options(exercise.timelimit = 60, exercise.checker = checker)
knitr::opts_chunk$set(error = TRUE, out.width = "100%")
```

```{r, context="server-start", include=FALSE}
# Capture metrics only if running an official primer hosted by RStudio
library(curl)
library(later)
tryCatch(
  source("https://metrics.rstudioprimers.com/learnr/installMetrics", local=TRUE), 
  error = function(e){ 
    print("Warning: An error occurred with the tracking code.")
  }
)
```

```{r, context = "render", results = 'asis', echo = FALSE}
tryCatch(
  source("https://metrics.rstudioprimers.com/learnr/installClient", local=TRUE)$value,
  error = function(e){ 
    print("Warning: An error occurred with the client code.")
  }
)
```

## && and ||

###   

This tutorial extends the Control Flow tutorial, where you learned how to use `if`, `else`, `return()`, and `stop()`.

Here you will learn how to 

1. combine logical tests in an if statement 
1. write if statements that work with vectors, which is a prerequisite if you want to write vectorized functions.

###  

Here's what `clean()` looked like at the end of the Control Flow tutorial. Do you notice that all of the if statements have the same outcome? 

```{r eval = FALSE}
clean <- function(x) {
  stopifnot(!is.null(x))
  if (x == -99) return(NA)
  if (x == ".") return(NA)
  if (x == "NaN") return(NA)
  x
}
```

Let's use your knowledge of logical tests to trim them down to a single if statement. 

###   

* Write a logical test that returns TRUE when x is -99 OR x is "." (Let's ignore the "NaN" case to keep things simple). Then click Submit Answer.

```{r logic, exercise = TRUE}

```

```{r logic-hint}
"You can combine two logical tests in R with `&` (and) and `|` (or), e.g. x < 0 & x > 1."
```

```{r logic-solution}
x == -99 | x == "."
```

```{r logic-check}
"This is the correct way to combine logical tests in R, but it has some downsides when you use it in an if statement."
```

### & and |

`&` and `|` are R's boolean operators for combining logical tests.

* `&` stands for "and" will return `TRUE` if _**both**_ tests return `TRUE` and will return `FALSE` otherwise. 
* `|` stands for "or" will return `TRUE` if _**one or both**_ tests returns `TRUE` and will return `FALSE` otherwise. 

So,

```{r}
x <- -99
x == -99 | x == "."
```

###  

However, it is bad practice to use `&` and `|` to combine logical tests within an `if` condition. Why? Because:

1. there is something better (as you'll see in a minute)
1. `&` and `|` tend to generate warning messages when used with `if`

###   

As R operators, both `&` and `|` are vectorized which means that you can use them with vectors. This is very useful.

```{r}
x <- c(-99, 0 , 1)
x == -99
x == "."
x == -99 | x == "."
```

###  

However, `if` conditions are not vectorized. `if` expects the logical test contained within its parentheses to return a **single** `TRUE` or `FALSE`. If the condition returns a vector of `TRUE` or `FALSE`s, `if` will use the first value and show a warning message.

```{r}
x <- c(-99, 0 , 1)
if (x == -99 | x == ".") NA
```

### && and ||

You can avoid this by always using `&&` and `||` within your `if` conditions. `&&` and `||` are lazy substitutes for `&` and `|`. They are lazy in two ways.

First, `&&` and `||` always return a single `TRUE` or `FALSE`. If you give `&&` or `||` vectors, they will compare only the first elements of the vectors---and they will not return a warning message.

```{r}
x <- c(-99, 0 , 1)
x == -99 || x == "."
```
    
### Use ||

Let's use this to our immediate advantage.

* Replace the two `if` statements below with a single statement that tests whether x is `-99` or `"."` without throwing error messages.

```{r or, exercise = TRUE}
clean <- function(x) {
  stopifnot(!is.null(x))
  if (x == -99) return(NA)
  if (x == ".") return(NA)
  x
}
```

```{r or-hint}
"Like |, || expects a _complete_ logical test on each side of ||."
```

```{r or-solution}
clean <- function(x) {
  stopifnot(!is.null(x))
  if (x == -99 || x == ".") return(NA)
  x
}
```

```{r or-check}
"Now lets see what happens if you use clean() with a vector of values."
```

###  

```{r eval = FALSE}
clean <- function(x) {
  stopifnot(!is.null(x))
  if (x == -99 || x == ".") return(NA)
  x
}
```

```{r vector-quiz, echo = FALSE}
question("What will `clean(c(-99, 0, 1))` return?",
         answer("NA 0 1", message = "The logical test will only consider the first value of the vector x, which here contains c(-99, 0, 1). Since that value is -99, the logical test will return TRUE and R will execute the code `return(NA)`, which returns a single NA."),
         answer("NA _with_ a warning message", message = "You certainly would get a warning message if your logical test used |, which will pass a vector of TRUEs and FALSEs to if, triggering the message. However, || will only pass on a single TRUE or FALSE."),
         answer("NA _without_ a warning message", correct = TRUE),
         answer("an error"),
         allow_retry = TRUE)
```

### Computation

The most important reason to use `||` instead of `|` is that `||` saves unnecessary computation when possible. This is the second way that `&&` and `||` are lazy.

When possible, `&&` and `||` jump to the correct conclusion after evaluating the first of the two logical tests (not so with `&` and `|`). 

* `&&` will return `FALSE` if the test on the left returns `FALSE` (because the combined test would return `FALSE`). 
* `||` will return `TRUE` if the test on the left returns `TRUE` (because the combined test would return `TRUE`)  
  
In either case, `&&` and `||` will not evaluate the test on the right.
    
```{r}
x <- -99
if (x == -99 || stop("if you evaluate this.")) "I didn't evaluate stop()."
```
 
###  

How could you use this?

Remember how this code returns an error because `if` cannot handle the result of `NULL == -99`?

```{r}
clean <- function(x) {
  if (x == -99) return(NA)
  x
}
clean(NULL)
```

### Quiz

Suppose we redefine `clean()` like this:

```{r eval = FALSE}
clean <- function(x) {
  if (is.null(x) || x == -99) return(NA)
  x
}
```

```{r null-quiz, echo = FALSE}
question("Now what will `clean(NULL)` return?",
         answer("An error", message = "clean() certainly would return an error if it evaluated `x == -99` (and collapsed the results into a single result to pass to `if`), but _does_ `clean()` evaluate `x == -99`?"),
         answer("NA", correct = TRUE, message = "|| first evaluates `is.null(x)`. `is.null(NULL)` returns `TRUE`, so `||` stops and returns `TRUE` to `if`. As a result, R never evaluates the code that leads to an error."),
         allow_retry = TRUE)
```

## Vectorized if

###  

Burried in the last section is an interesting question: what if you _do_ want `clean()` to work with vectors? i.e.

```{r echo = FALSE}
clean <- function(x) ifelse(x == -99, NA, x)
```

```{r}
clean(c(-99, 0, 1))
```

That would be a handy way to clean whole columns of data. How could you do it?

###   

Compare these two functions (one should seem familiar). What is different?

```{r}
clean <- function(x) {
  if (x == -99) NA else x
}

clean2 <- function(x) {
  ifelse(x == -99, NA, x)
}
```

```{r ifelse-quiz, echo = FALSE}
question("Make a guess: do these two functions do the same thing?",
         answer("Yes!", correct = TRUE, message = "Well, you are sort of correct..."),
         answer("No!", correct = TRUE, message = "Well, you are sort of correct..."),
         answer("Sort of...", correct = TRUE, message = "You are absolutely right."), 
         allow_retry = TRUE,
         type = "single")
```

### ifelse()

`ifelse()` is a function that replicates an if else statement. It takes three arguments: a logical test followed by two pieces of code. If the test returns `TRUE`, `ifelse()` will return the results of the first piece of code. If the test returns `FALSE`, `ifelse()` will return the results of the second piece of code. 

So `clean(-99)` and `clean2(-99)` both return `NA`.

```{r}
clean(-99)
clean2(-99)
```

###  

However, unlike `if` and `else`, `ifelse` is vectorized. As a result, you can pass `ifelse()` a vector of values and it will apply the implied if else statement separately to each element of the vector.

```{r}
x <- c(-99, 0, 1)
ifelse(x == -99, NA, x)
```

`clean2()` inherits this vectorized property from `ifelse()`. 

```{r}
clean2(c(-99, 0, 1))
```

Compare that to `clean()` (which is non-vectorized because it relies on `if` and `else`, which are non-vectorized).

```{r}
clean(c(-99, 0, 1))
```

### if_else

The dplyr package offers a slight improvement on `ifelse()` named `if_else()`. `if_else()` is faster than `ifelse()`, but it requires you to make sure that each case in the if else statement returns the same type of object. For example, the statement needs to return a real number (or a string, or a logical, etc.) _whether or not_ the condition is `TRUE`.

No big deal, right? Well kind of.

```{r}
x <- c(-99, 0, 1)
if_else(x == -99, NA, x)
```

### NA

What happened? Recall that data in R comes in six atomic types. 

```{r type-quiz, echo = FALSE}
question("What type of data is NA?",
         answer("double (i.e. real numbers)"),
         answer("integer"),
         answer("character"),
         answer("logical", correct = TRUE, message = "It turns out that if you type NA at the command line, R saves the result as a logical."),
         answer("complex"),
         answer("raw"),
         allow_retry = TRUE)
```

###  

It is true:

```{r}
typeof(NA)
```

So when you write `if_else(x == -99, NA, x)`, `if_else()` returns a logical in the first case and a double (real number) in the second (assuming `x` is a real number). 

You can get around this mishap in two ways:

1. Stick to `ifelse()`
1. Use a NA that comes with a type

### Types of NA

You may not realize it, but R comes with five types of NA. They all appear as `NA` when printed, but they are each saved with a separate data type. These are:

```{r}
NA # logical
NA_integer_ # integer
NA_real_ # double
NA_complex_ # complex
NA_character_ # character
```

You can fix `if_else()` by being precise about which NA to use (most other R functions will convert the type of NA without bothering you).

```{r}
x <- c(-99, 0, 1)
if_else(x == -99, NA_real_, x)
```

### Use if_else

* Fix the `if_else()` statement of `clean2()` to work with real numbers. Then click Submoit Answer.

```{r cleans, exercise = TRUE}
clean2 <- function(x) {
  ifelse(x == -99, NA, x)
}
```

```{r cleans-solution}
clean2 <- function(x) {
  ifelse(x == -99, NA_real_, x)
}
```

```{r cleans-check}
"Notice that this version of `clean2()` will work with real numbers, but not other types of data like characters. What if you want `clean2()` to work with all types of data? That's simple: stick with `ifelse()`."
```

## Vectorized else if

###  

What if you want to write a vectorized version of a multi-part if else tree? Like the tree in this function:

```{r eval = FALSE}
clean <- function(x) {
  if (x == -99) NA 
  else if (x == ".") NA
  else if (x == "") NA
  else if (x == "NaN") NA
  else x
}
```

In this case, neither `ifelse()` or `if_else()` will do. Why? Because each can only handle a single if condition, but our tree has four.

### case_when()

You can vectorize multi-part if else statements with dplyr's `case_when()` function. Here is how you would use `case_when()` to rewrite our `foo()` function from the Control Flow tutorial.

Here is the masterpiece in its original form

```{r}
foo <- function(x) {
  if (x > 2) "a"
  else if (x < 2) "b"
  else if (x == 1) "c"
  else "d"
}
```

And here it is with `case_when()`.

```{r}
foo2 <- function(x) {
  case_when(
    x > 2  ~ "a",
    x < 2  ~ "b",
    x == 1 ~ "c",
    TRUE   ~ "d"
  )
}
```

###  

And here are our foos in action to prove that `foo2()` is vectorized.

```{r}
x <- c(3, 2, 1)
foo(x)
foo2(x)
```

Notice that 

1. `case_when()` returns a single case for each element, the first case whose left hand side evaluates to `TRUE`
1. The left hand side of the last case evaluates to `TRUE` no matter what the value of `x` is (In fact, the left hand side _is_ `TRUE`). This is an easy way to add an `else` clause to the end of `case_when()`.

Now let's look at the unusual syntax of `case_when()`.

### case_when() syntax

```{r eval = FALSE}
foo2 <- function(x) {
  case_when(
    x > 2  ~ "a",
    x < 2  ~ "b",
    x == 1 ~ "c",
    TRUE   ~ "d"
  )
}
```

Each argument of `case_when()` is a pair that consists of a logical test on the left hand side and a piece of code on the right hand side. The two are _always_ separated by a `~`.

Like `if_else()`, `case_when()` expects each case to return the same type of output. So keep those NA types handy: `NA`, `NA_integer_`, `NA_real_`, `NA_complex_`, `NA_character_`.

### Final Challenge

* Rewrite the multi-part version of `clean()` to use `case_when()`, which will allow `clean()` to handle vectors. Retain each case. Assume where necessary that `clean()` will only work with real numbers. Then click Submit Answer.

```{r when, exercise = TRUE}
clean <- function(x) {
  if (x == -99) NA 
  else if (x == ".") NA
  else if (x == "") NA
  else if (x == "NaN") NA
  else x
}
```

```{r when-hint}
"Use NA's that have the right type."
```

```{r when-solution}
clean <- function(x) {
  case_when(
    x == -99 ~ NA_real_, 
    x == "." ~ NA_real_,
    x == "" ~ NA_real_,
    x == "NaN" ~ NA_real_,
    TRUE ~ x
  )
}
```

```{r when-check}
'And if you noticed that a vector of real numbers would never contain ".", "", and "Nan" because they are strings, you are of course right. Thanks for playing along with the charade.'
```

### Congratulations!

You've learned how to alter the control flow of your functions with:

* `if`
* `else`
* `return()`
* `stop()`
* `stopifnot()`
* `ifelse()`

Not only that, you tackled two advanced methods: dplyr's `if_else()` and dplyr's `case_when()`.


