---
title: "Customize plots"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
tutorial:
  id: "com.rstudio.primers.08-customize"
  version: 1.0
---

```{r setup, include=FALSE}
library(learnr)
library(mgcv)
library(tidyverse)
library(ggthemes)
library(RColorBrewer)
library(viridis)

p <- ggplot(diamonds) +
  geom_boxplot(mapping = aes(x = cut, y = price))

p1 <- ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_smooth(mapping = aes(color = cut), se = FALSE) + 
  labs(title = "Carat size vs. Price",
       subtitle = "Fair cut diamonds tend to be large, but they fetch the lowest prices for most carat sizes.",
       caption = "Data by Hadley Wickham")

p_cont <- ggplot(data = mpg) + 
  geom_jitter(mapping = aes(x = displ, y = hwy, color = hwy)) +
  theme_bw()

p_legend <- ggplot(data = mpg) + 
  geom_jitter(mapping = aes(x = displ, y = hwy, color = class, fill = hwy), 
              shape = 21, size = 3, stroke = 1) +
  theme_bw()

checker <- function(label, user_code, check_code, envir_result, evaluate_result, ...) {
  list(message = check_code, correct = TRUE, location = "append")
}
tutorial_options(exercise.timelimit = 60, exercise.checker = checker)
knitr::opts_chunk$set(error = TRUE, out.width = "100%", message = FALSE)
```

```{r, context="server-start", include=FALSE}
# Capture metrics only if running an official primer hosted by RStudio
library(curl)
library(later)
tryCatch(
  source("https://metrics.rstudioprimers.com/learnr/installMetrics", local=TRUE), 
  error = function(e){ 
    print("Warning: An error occurred with the tracking code.")
  }
)
```

```{r, context = "render", results = 'asis', echo = FALSE}
tryCatch(
  source("https://metrics.rstudioprimers.com/learnr/installClient", local=TRUE)$value,
  error = function(e){ 
    print("Warning: An error occurred with the client code.")
  }
)
```

## Welcome

This tutorial will teach you how to customize the look and feel of your plots. You will learn how to:

* **Zoom in** on areas of interest
* Add **labels** and **annotations** to your plots
* Change the appearance of your plot with a **theme**
* Use **scales** to select custom color palettes
* Modify the labels, title, and position of **legends**

The tutorial is adapted from _R for Data Science_ by Hadley Wickham and Garrett Grolemund, published by O’Reilly Media, Inc., 2016, ISBN: 9781491910399. You can purchase the book at [shop.oreilly.com](http://shop.oreilly.com/product/0636920034407.do).

The tutorial uses the ggplot2, dplyr, scales, ggthemes, and viridis packages, which have been pre-loaded for your convenience.

## Zooming

In the previous tutorials, you learned how to visualize data with graphs. Now let's look at how to customize the look and feel of your graphs. To do that we will need to begin with a graph that we can customize.

### Review 1 - Make a plot

In the chunk below, make a plot that uses boxplots to display the relationship between the `cut` and `price` variables from the diamonds dataset.

```{r r1, exercise = TRUE}

```

```{r r1-solution}
ggplot(diamonds) +
  geom_boxplot(mapping = aes(x = cut, y = price))
```

```{r r1-check}
"Good job! Let's use this plot as a starting point to make a more pleasing plot that displays a clear message."
```

### Storing plots

Since we want to use this plot again later, let's go ahead and save it.

```{r}
p <- ggplot(diamonds) +
  geom_boxplot(mapping = aes(x = cut, y = price))
```

Now whenever you call `p`, R will draw your plot. Try it and see.

```{r r2, exercise = TRUE}

```

```{r r2-solution}
p
```

```{r r2-check}
"Good job! By the way, have you taken a moment to look at what the plot shows? Let's do that now."
```

### Surprise?

Our plot shows something surprising: when you group diamonds by cut, the worst cut diamonds have the highest median price. It's a little hard to see in the plot, but you can verify it with some data manipulation.

```{r}
diamonds %>% 
  group_by(cut) %>% 
  summarise(median = median(price))
```

### Zoom

```{r echo = FALSE, out.width = "80%"}
p
```

The difference between median prices is hard to see in our plot because each group contains distant outliers. 

We can make the difference easier to see by zooming in on the low values of $y$, where the medians are located. There are two ways to zoom with ggplot2: with and without clipping.

### Clipping

Clipping refers to how R should treat the data that falls outside of the zoomed region. To see its effect, look at these plots. Each zooms in on the region where price is between $0 and \$7,500. 

```{r echo = FALSE, out.width = "50%", warning = FALSE, message = FALSE}
p + ylim(0, 7500)
p + coord_cartesian(ylim = c(0, 7500))
```

* The plot on the left zooms _by_ clipping. It removes all of the data points that fall outside of the desired region, and then plots the data points that remain.
* The plot on the right zooms _without_ clipping. You can think of it as drawing the entire graph and then zooming into a certain region. 

### xlim() and ylim()

Of these, zooming by clipping is the easiest to do. To zoom your graph on the $x$ axis, add the function `xlim()` to the plot call. To zoom on the $y$ axis add the function `ylim()`. Each takes a minimum value and a maximum value to zoom to, like this

```{r eval = FALSE}
some_plot +
  xlim(0, 100)
```

### Exercise 1 - Clipping

Use `ylim()` to recreate our plot on the left from above. The plot zooms the $y$ axis from 0 to 7,500 by clipping.

```{r r3, exercise = TRUE, warning = FALSE}
p
```

```{r r3-solution}
p + ylim(0, 7500)
```

```{r r3-check}
"Good job! Zooming by clipping will sometimes make the graph you want, but in our case it is a very bad idea. Can you tell why?"
```

### A caution

Zooming by clipping is a bad idea for boxplots. `ylim()` fundamentally changes the information conveyed in the boxplots because it throws out some of the data before drawing the boxplots. Those aren't the medians of the entire data set that we are looking at.

How then can we zoom without clipping?

### xlim and ylim

To zoom without clipping, set the xlim and/or ylim arguments of your plot's `coord_` function. Each takes a numeric vector of length two (the minimum and maximum values to zoom to). 

This is easy to do if your plot explicitly calls a `coord_` function

```{r out.width = "80%"}
p + coord_flip(ylim = c(0, 7500))
```

### coord_cartesian()

But what if your plot doesn't call a `coord_` function? Then your plot is using Cartesian coordinates (the default). You can adjust the limits of your plot without changing the default coordinate system by adding `coord_cartesian()` to your plot.

Try it below. Use `coord_cartesian()` to zoom `p` to the region where price falls between 0 and 7500.

```{r r4, exercise = TRUE}

```

```{r r4-solution}
p + coord_cartesian(ylim = c(0, 7500))
```

```{r r4-check}
"Good job! Now it is much easier to see the differences in the median."
```

### p

Notice that our code so far has used `p` to make a plot, but it hasn't changed the plot that is saved inside of `p`. You can run `p` by itself to get the unzoomed plot.

```{r out.width = "80%"}
p
```

### Updating p

I like the zooming, so I'm purposefully going to overwrite the plot stored in `p` so that it uses it.

```{r out.width = "80%"}
p <- p + coord_cartesian(ylim = c(0, 7500))
p
```

## Labels

### labs()

The relationship in our plot is now easier to see, but that doesn't mean that everyone who sees our plot will spot it. We can draw their attention to the relationship with a label, like a title or a caption.

To do this, we will use the `labs()` function. You can think of `labs()` as an all purpose function for adding labels to a ggplot2 plot.

### Titles

Give `labs()` a title argument to add a title.

```{r out.width = "80%"}
p + labs(title = "The title appears here")
```

### Subtitles

Give `labs()` a subtitle argument to add a subtitle. If you use multiple arguments, remember to separate them with a comma.

```{r out.width = "80%"}
p + labs(title = "The title appears here",
         subtitle = "The subtitle appears here, slightly smaller")
```

### Captions

Give `labs()` a caption argument to add a caption. I like to use captions to cite my data source.

```{r out.width = "80%"}
p + labs(title = "The title appears here",
         subtitle = "The subtitle appears here, slightly smaller",
         caption = "Captions appear at the bottom.")
```

### Exercise 2 - Labels

Plot `p` with a set of informative labels. for learning purposes, be sure to use a title, subtitle, and caption.

```{r r5-setup}
p <- ggplot(diamonds) +
  geom_boxplot(mapping = aes(x = cut, y = price)) + 
  coord_cartesian(ylim = c(0, 7500))
```


```{r r5, exercise = TRUE}

```

```{r r5-solution}
p + labs(title = "Diamond prices by cut",
         subtitle = "Fair cut diamonds fetch the highest median price. Why?",
         caption = "Data collected by Hadley Wickham")
```

```{r r5-check}
"Good job! By the way, why *do* fair cut diamonds fetch the highest price?"
```

### Exercise 3 - Carat size? 

Perhaps a diamond's cut is conflated with its carat size. If fair cut diamonds tend to be larger diamonds that would explain their larger prices. Let's test this.

Make a plot that displays the relationship between carat size, price, and cut for all diamonds. How do you interpret the results? Give your plot a title, subtitle, and caption that explain the plot and convey your conclusions.

If you are looking for a way to start, I recommend using a smooth line with color mapped to cut, perhaps overlaid on the background data.

```{r r6, exercise = TRUE}

```

```{r r6-solution}
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_smooth(mapping = aes(color = cut), se = FALSE) + 
  labs(title = "Carat size vs. Price",
       subtitle = "Fair cut diamonds tend to be large, but they fetch the lowest prices for most carat sizes.",
       caption = "Data by Hadley Wickham")
```

```{r r6-check}
"Good job! The plot corroborates our hypothesis."
```

### p1

Unlike `p`, our new plot uses color and has a legend. Let's save it to use later when we learn to customize colors and legends.

```{r out.width = "80%"}
p1 <- ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_smooth(mapping = aes(color = cut), se = FALSE) + 
  labs(title = "Carat size vs. Price",
       subtitle = "Fair cut diamonds tend to be large, but they fetch the lowest prices for most carat sizes.",
       caption = "Data by Hadley Wickham")
```

### annotate()

`annotate()` provides a final way to label your graph: it adds a single geom to your plot. When you use `annotate()`, you must first choose which type of geom to add. Next, you must manually supply a value for each aesthetic required by the geom.

So for example, we could use `annotate()` to add text to our plot. 

```{r}
p1 + annotate("text", x = 4, y = 7500, label = "There are no cheap,\nlarge diamonds")
```

Notice that I select `geom_text()` with `"text"`, the suffix of the function name in quotation marks.

In practice, I find `annotate()` time consuming to work with, but you can accomplish quite a lot with `annotate()` if you take the time.

## Themes 

One of the most effective ways to control the look of your plot is with a theme.

### What is a theme?

A theme describes how the non-data elements of your plot should look. For example, these two plots show the same data, but they use two very different themes.

```{r echo = FALSE, out.width ="50%", message = FALSE, warning = FALSE}
p1 + theme_bw()
p1 + theme_economist()
```

### Theme functions

To change the theme of your plot, add a `theme_` function to your plot call. The ggplot2 package provides eight theme functions to choose from.

* `theme_bw()`
* `theme_classic()`
* `theme_dark()`
* `theme_gray()`
* `theme_light()`
* `theme_linedraw()`
* `theme_minimal()`
* `theme_void()`

Use the box below to plot `p1` with each of the themes. Which theme do you prefer? Which theme does ggplot2 apply by default?

```{r r7, exercise = TRUE}
p1 + theme_bw()
```

```{r r7-solution}
p1 + theme_classic()
```

```{r r7-check}
"Good Job! ggplot2 uses theme_gray()` by default."
```

### ggthemes

If you would like to give your graph a more complete makeover, the **ggthemes** package provides extra themes that imitate the graph styles of popular software packages and publications. These include:

* `theme_base()`
* `theme_calc()`
* `theme_economist()`
* `theme_economist_white()`
* `theme_excel()`
* `theme_few()`
* `theme_fivethirtyeight()`
* `theme_foundation()`
* `theme_gdocs()`
* `theme_hc()`
* `theme_igray()`
* `theme_map()`
* `theme_pander()`
* `theme_par()`
* `theme_solarized()`
* `theme_solarized_2()`
* `theme_solid()`
* `theme_stata()`
* `theme_tufte()`
* `theme_wsj()`

Try plotting `p1` with at least two or three of the themes mentioned above.

```{r r8, exercise = TRUE}
p1
```

```{r r8-solution}
p1 + theme_wsj()
```

```{r r8-check}
"Good Job! Notice that each theme supplies its own font sizes, which means that your captions might run off the page for some themes. In practice, you can fix this by resizing your graph window."
```

### Update p1

If you compare the ggtheme themes to the styles they imitate, you might notice something: the colors used to plot your data haven't changed. The colors are noticeably ggplot2 colors. In the next section, we'll look at how to customize this remaining part of your graph: the data elements.

Before we go on, I suggest that we update `p1` to use `theme_bw()`. It will make our next set of modifications easier to see.

```{r p1, out.width = "80%", messages = FALSE}
p1 <- p1 + theme_bw()
p1
```

## Scales

### What is a scale?

Every time you map an aesthetic to a variable, ggplot2 relies on a _scale_ to select the specific colors, sizes, or shapes to use for the values of your variable. 

A scale is an R function that works like a mathematical function; it maps each value in a data space to a level in an aesthetic space. But it may be easier to think of a scale as a "palette." When you give your graph a color scale, you give it a palette of colors to use.

### Using scales

ggplot2 chooses a pleasing set of scales to use whenever you make a graph. You can change or customize these scales by adding a scale function to your plot call. 

For example, the code below plots `p1` in greyscale instead of the default colors.

```{r}
p1 + scale_color_grey()
```

### A second example

You can add scales for every aesthetic mapping, including the $x$ and $y$ mappings (the code below log transforms the x and y axes). 

```{r}
p1 +
  scale_x_log10() + 
  scale_y_log10()
```

ggplot2 supplies over 50 scales to use. This may seem overwhelming, but the scales are organized according to an intuitive naming convention.

### Naming convention

ggplot2 scale functions follow a naming convention. Each function name contains the same three elements in order, separated by underscores:

* The prefix `scale` 
* the name of an aesthetic, which the scale adjusts (e.g. `color`, `fill`, `size`)
* a unique label for the scale (e.g. `grey`, `brewer`, `manual`)

`scale_shape_manual()` and `scale_x_continuous()` are examples of the naming scheme.

You can see the complete list of scale names at [http://ggplot2.tidyverse.org/reference/](http://ggplot2.tidyverse.org/reference/). In this tutorial, we will focus on scales that work with the color aesthetic.

### Discrete vs. continuous

Scales specialize in either _discrete_ variables or _continuous_ variables. In other words, you would use a different set of scales to map a discrete variable, like diamond clarity, than you would use to map a continuous variable, like diamond price.

```{r q1, echo = FALSE}
question("Which type of variable does `p1` map to the color aesthetic?",
         answer("Discrete", correct = TRUE, message = "p1 maps color to cut, a discrete variable with five distinct levels."),
         answer("Continuous"),
         allow_retry = TRUE)
```

### scale_color_brewer

One of the most useful color palettes for discrete variables is `scale_color_brewer()` (`scale_fill_brewer()` if you are working with fill. Run the code below to see the effect of the scale.

```{r make_p1, include = FALSE}
p1 <- ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_smooth(mapping = aes(color = cut), se = FALSE) + 
  labs(title = "Carat size vs. Price",
       subtitle = "Fair cut diamonds tend to be large, but they fetch the lowest prices for most carat sizes.",
       caption = "Data by Hadley Wickham") + theme_bw()
```

```{r r9, exercise = TRUE, warning = FALSE, message = FALSE, exercise.setup="make_p1"}
p1 + scale_color_brewer()
```

```{r r9-check}
"Good job! scale_color_brewer() applies a color palette from the RColorBrewer package, a package that specializes in attractive color palettes." 
```

### RColorBrewer

The RColorBrewer package contains a variety of palettes developed by Cynthis Brewer. Each palette is designed to look pleasing as well as to differentiate between the values represented by the palette. You can learn more about the color brewer project at [colorbrewer2.org](http://colorbrewer2.org).

Altogether, the RColorBrewer package contains 35 palettes. You can see each palette and its name by running `RColorBrewer::display.brewer.all()`. Try it below.

```{r r10, exercise = TRUE, warning = FALSE, message = FALSE, fig.width=4, fig.height = 7}

```

```{r r10-check}
"Good job! Our graph above used the Blues palette (the default)." 
```

```{r r10-solution}
RColorBrewer::display.brewer.all()
```

### Brewer palettes

By default, `scale_color_brewer()` will use the "Blues" palette from the RColorBrewer package. To use a different RColorBrewer palette, set the palette argument of `scale_color_brewer()` to one of the RColorBrewer palette names, surrounded by quotation marks, e.g.

```{r warnings = FALSE, out.width = "80%"}
p1 + scale_color_brewer(palette = "Purples")
```

### Exercise - scale_color_brewer()

Recreate the graph below, which uses a different palette from the RColorBrewer package.

```{r echo = FALSE, out.width = "80%"}
p1 + scale_color_brewer(palette = "Spectral")
```

```{r r11, exercise = TRUE, warning = FALSE, message = FALSE, exercise.setup="make_p1"}


```

```{r r11-solution}
p1 + scale_color_brewer(palette = "Spectral")
```

```{r r11-check}
"Good job! scale_color_brewer() is one of the most useful functions for customizing colors in ggplot2 because it does for you the hard work of selecting a pleasing combination of colors. If you'd like to select individual colors yourself, try the scale_color_manual() function."
```

### Continuous colors

`scale_color_brewer()` works with discrete variables, but what if your plot maps color to a continuous variable?

Since we do not have a plot that applies color to a continuous variable, let's make one. 

```{r}
p_cont <- ggplot(data = mpg) + 
  geom_jitter(mapping = aes(x = displ, y = hwy, color = hwy)) +
  theme_bw()

p_cont
```

### Discrete vs. continuous in action

If we apply `scale_color_brewer()` to our new plot, we get an error message that confirms what you know: you cannot use a scale that is built for discrete variables to customize the mapping to a continuous variable.

```{r}
p_cont + scale_color_brewer()
```

### distiller

Luckily, `scale_color_brewer()` has a comes with a continuous analogue named `scale_color_distiller()` (also `scale_fill_distiller()`). 

Use `scale_color_distiller()` just as you would `scale_color_brewer()`. `scale_color_distiller()` will take any RColorBrewer palette, and interpolate between colors as necessary to provide an entire continuous range of colors.

So for example, we could reuse the Spectral palette in our continuous plot

```{r}
p_cont + scale_color_distiller(palette = "Spectral")
```

### Exercise - scale_color_distiller()

Recreate the graph below, which uses a different palette from the RColorBrewer package.

```{r echo = FALSE, out.width = "80%"}
p_cont + scale_color_distiller(palette = "BrBG")
```

```{r r12, exercise = TRUE, warning = FALSE, message = FALSE, exercise.setup="make_p1"}

```

```{r r12-solution}
p_cont + scale_color_distiller(palette = "BrBG")
```

```{r r12-check}
"Good job! ggplot2 also supplies scale_color_gradient(), scale_color_gradient2(), and scale_color_gradientn(), which you can use to  construct gradients manually between 2, 3, and n colors."
```

### viridis

The viridis package contains a collection of very good looking color palettes for continuous variables. Each palette is designed to show the gradation of continuous values in an attractive, and perceptionally uniform way (no range of values appears more important than another). As a bonus, the palettes are both color blind and black and white printer friendly!

To add a viridis palette, use `scale_color_viridis()` or `scale_fill_viridis()`, both of which come in the viridis package.

```{r}
p_cont + scale_color_viridis()
```

### viridis options

Altogether, the viridis package comes with four color palettes, named magma, plasma, inferno, and viridis.

However, you do not select the palettes by name. To select a viridis color palette, set the `option` argument of `scale_color_viridis()` to one of `"A"` (magma), `"B"` (plasma), `"C"` (inferno), or `"D"` (viridis). 

Try each option with `p_cont` below. Determine which is the default.

```{r r13, exercise = TRUE, warning = FALSE, message = FALSE, exercise.setup="make_p1"}

```

```{r r13-solution}
p_cont + scale_color_viridis(option = "D")
```

```{r r13-check}
"Good job! Option D is the default if you do not select an option."
```

## Legends

### Customizing a legend

The last piece of a ggplot2 graph to customize is the legend. When it comes to legends, you can customize the:

* position of the legend within the graph
* the "type" of the legend, or whether a legend appears at all
* the title and labels in the legend

Customizing legends is a little more chaotic than customizing other parts of the graph, because the information that appears in a legend comes from several different places.

### Positions

To change the position of a legend in a ggplot2 graph add one of the below to your plot call:

* `+ theme(legend.position = "bottom")`
* `+ theme(legend.position = "top")`
* `+ theme(legend.position = "left")`
* `+ theme(legend.position = "right")` (the default)

Try this now. Move the legend in `p_cont` to the bottom of the graph.

```{r r14, exercise = TRUE}

```

```{r r14-solution}
p_cont + theme(legend.position = "bottom")
```

```{r r14-check}
"Good job! If you move the legend to the top or bottom of the plot, ggplot2 will reogranize the orientation of the legend from vertical to horizontal."
```

### theme() vs. themes

Theme functions like `theme_grey()` and `theme_bw()` also adjust the legend position (among all of the other details they orchestrate). So if you use `theme(legend.position = "bottom")` in your plots, be sure to add it _after_ any `theme_` functions you call, like this

```{r r15, exercise = TRUE, exercise.eval = TRUE, out.width = "80%"}
ggplot(data = mpg) + 
  geom_jitter(mapping = aes(x = displ, y = hwy, color = hwy)) +
  theme_bw() +
  theme(legend.position = "bottom")
```

If you do this, ggplot2 will apply all of the settings of `theme_bw()`, and then overwrite the legend position setting to "bottom" (instead of vice versa).

### Types

You may have noticed that color and fill legends take two forms. If you map color (or fill) to a discrete variable, the legend will look like a standard legend. This is the case for the bottom legend below. 

If you map color (or fill) to a continuous legend, your legend will look like a colorbar. This is the case in the top legend below. The color bar helps convey the continuous nature of the variable.

```{r plegend}
p_legend <- ggplot(data = mpg) + 
  geom_jitter(mapping = aes(x = displ, y = hwy, color = class, fill = hwy), 
              shape = 21, size = 3, stroke = 1) +
  theme_bw()

p_legend
```

### Changing type

You can use the `guides()` function to change the type or presence of each legend in the plot. To use `guides()`, type the name of the aesthetic whose legend you want to alter _as and argument name_. Then set it to one of

* `"legend"` - to force a legend to appear as a standard legend instead of a colorbar
* `"colorbar"` - to force a legend to appear as a colorbar instead of a standard legend. Note: this can only be used when the legend can be printed as a colorbar (in which case the default will be colorbar).
* `"none"` - to remove the legend entirely. This is useful when you have redundant aesthetic mappings, but it may make your plot indecipherable otherwise.

```{r out.width = "80%"}
p_legend + guides(fill = "legend", color = "none")
```

### Exercise - guides()

Use `guides()` to remove each legend from the `p_legend` plot.

```{r r16, exercise = TRUE, exercise.setup = "plegend"}

```

```{r r16-solution}
p_legend + guides(fill = "none", color = "none")
```

```{r r16-check}
"Good job! If you move the legend to the top or bottom of the plot, ggplot2 will reogranize the orientation of the legend from vertical to horizontal."
```

### Labels

To control the title and labels of a legend, you must turn to the `scale_` functions. Each `scale_` function takes a name and a labels argument, which it will use to build the legend associated with the scale. The labels argument should be a vector of strings that has one string for each label in the default legend.

So for example, you can adjust the legend of p1 with

```{r out.width = "80%"}
p1 + scale_color_brewer(name = "Cut Grade", labels = c("Very Bad", "Bad", "Mediocre", "Nice", "Very Nice"))
```

### What if?

This is handy, but it raises a question: what if you haven't invoked a `scale_` function to pass labels to? For example, the graph below relies on the default scales.

```{r echo = FALSE, out.width = "80%"}
p1
```

### Default scales

In this case, you need to identify the default scale used by the plot and then manually add that scale to the plot, setting the labels as you do. 

For example, our plot above relies on the default color scale for a discrete variable, which happens to be `scale_color_discrete()`. If you know this, you can relabel the legend like so:

```{r out.width = "80%"}
p1 + scale_color_discrete(name = "Cut Grade", labels = c("Very Bad", "Bad", "Mediocre", "Nice", "Very Nice"))
```


### Scale defaults

As you can see, it is handy to know which scales a ggplot2 graph will use by default. Here's a short list.

aesthetic	| variable   |	default
--------- | ---------- | -------------------------
x	        | continuous |	scale_x_continuous()
	        | discrete   |	scale_x_discrete()
y	        | continuous |	scale_y_continuous()
	        | discrete	 |  scale_y_discrete()
color	    | continuous |	scale_color_continuous()
	        | discrete	 |  scale_color_discrete()
fill	    | continuous |	scale_fill_continuous()
	        | discrete   |	scale_fill_discrete()
size	    | continuous |	scale_size()
shape	    | discrete	 | scale_shape()

### Exercise - Legends

Use the list of scale defaults above to relabel the legend in `p_cont`. The legend should have the title "Highway MPG". Also place the legend at the top of the plot.

```{r r17, exercise = TRUE}
p_cont
```

```{r r17-solution}
p_cont + scale_color_continuous(name = "Highway MPG") + theme(legend.position = "top")
```

```{r r17-check}
"Good job!"
```


### Axis labels

In ggplot2, the axes are the "legends" associated with the $x$ and $y$ aesthetics. As a result, you can control axes titles and labels in the same way as you control legend titles and labels:

```{r}
p1 + scale_x_continuous(name = "Carat Size", labels = c("Zero", "One", "Two", "Three", "Four", "Five"))
```

## Quiz

In this tutorial, you learned how to customize the graphs that you make with ggplot2 in several ways. You learned how to:

* Zoom in on regions of the graph
* Add titles, subtitles, and annotations
* Add themes
* Add color scales
* Adjust legends

To cement your skills, combine what you've learned to recreate the plot below.

```{r echo = FALSE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point() + 
  geom_smooth(aes(color = cut), se = FALSE) +  
  labs(title = "Ideal cut diamonds command the best price for every carat size",
       subtitle = "Lines show GAM estimate of mean values for each level of cut",
       caption = "Data provided by Hadley Wickham",
       x = "Log Carat Size",
       y = "Log Price Size",
       color = "Cut Rating") +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_brewer(palette = "Greens") +
  theme_light()
```

```{r r18, exercise = TRUE}

```

```{r r18-solution}
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point() + 
  geom_smooth(mapping = aes(color = cut), se = FALSE) +  
  labs(title = "Ideal cut diamonds command the best price for every carat size",
       subtitle = "Lines show GAM estimate of mean values for each level of cut",
       caption = "Data provided by Hadley Wickham",
       x = "Log Carat Size",
       y = "Log Price Size",
       color = "Cut Rating") +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_brewer(palette = "Greens") +
  theme_light()
```



