---
title: "Scatterplots"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
tutorial:
  id: "com.rstudio.primers.05-scatterplots"
  version: 1.0
---

```{r setup, include=FALSE}
library(learnr)
library(ggrepel)
library(tidyverse)

checker <- function(label, user_code, check_code, envir_result, evaluate_result, ...) {
  list(message = check_code, correct = TRUE, location = "append")
}
tutorial_options(exercise.timelimit = 60, exercise.checker = checker)
knitr::opts_chunk$set(error = TRUE, out.width = "100%")
```

```{r, context="server-start", include=FALSE}
# Capture metrics only if running an official primer hosted by RStudio
library(curl)
library(later)
tryCatch(
  source("https://metrics.rstudioprimers.com/learnr/installMetrics", local=TRUE), 
  error = function(e){ 
    print("Warning: An error occurred with the tracking code.")
  }
)
```

```{r, context = "render", results = 'asis', echo = FALSE}
tryCatch(
  source("https://metrics.rstudioprimers.com/learnr/installClient", local=TRUE)$value,
  error = function(e){ 
    print("Warning: An error occurred with the client code.")
  }
)
```

## Welcome 

A **scatterplot** displays the relationship between two continuous variables. Scatterplots are one of the most common types of graphs---in fact, you've met scatterplots already in [Visualization Basics](https://tutorials.shinyapps.io/02-Vis-Basics/). 

In this tutorial, you'll learn how to:

* Make new types of scatterplots with `geom_text()` and `geom_jitter()`
* Add multiple **layers** of geoms to a plot
* Enhance scatterplots with `geom_smooth()`, `geom_rug()`, and `geom_repel()`
* Change the **coordinate system** of a plot

The tutorial is adapted from _R for Data Science_ by Hadley Wickham and Garrett Grolemund, published by O’Reilly Media, Inc., 2016, ISBN: 9781491910399. You can purchase the book at [shop.oreilly.com](http://shop.oreilly.com/product/0636920034407.do).

The tutorial uses the ggplot2, ggrepel, and dplyr packages, which have been pre-loaded for your convenience.

## Scatterplots

### Review 1 - geom_point()

In [Visualization Basics](https://tutorials.shinyapps.io/02-Vis-Basics/), you learned how to make a scatterplot with `geom_point()`.

The code below summarises the mpg data set and begins to plot the results. Finish the plot with `geom_point()`. Put `mean_cty` on the $x$ axis and `mean_hwy` on the $y$ axis.

```{r r1, exercise = TRUE}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot()
```

```{r r1-solution}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot() +
    geom_point(mapping = aes(x = mean_cty, y = mean_hwy))
```

```{r r1-check}
"Good job! It can be tricky to remember when to use %>% and when to use +. Use %>% to add one complete step to a pipe of code. Use + to add one more line to a ggplot2 call."
```


### geom_text()

`geom_text()` and `geom_label()` create scatterplots that use words instead of points to display data. Each requires the extra aesthetic `label`, which you should map to a variable that contains text to display for each observation.

Convert the plot below from `geom_point()` to `geom_text()` and map the label aesthetic to the class variable. When you are finished convert the code to `geom_label()` and rerun the plot. Can you spot the difference?

```{r r2, exercise = TRUE, exercise.eval = TRUE}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot() +
    geom_point(mapping = aes(x = mean_cty, y = mean_hwy))
```

```{r r2-solution}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot() +
    geom_text(mapping = aes(x = mean_cty, y = mean_hwy, label = class))

mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot() +
    geom_label(mapping = aes(x = mean_cty, y = mean_hwy, label = class))
```

```{r r2-check}
"Good job! geom_text() replaces each point with a piece of text supplied by the label aesthetic. geom_label replaces each point with a textbox. Notice that some pieces of text overlap each other, and others run off the page. We'll soon look at a way to fix this."
```


### geom_smooth()

In [Visualization Basics](https://tutorials.shinyapps.io/02-Vis-Basics/), you met `geom_smooth()`, which provides a summarised version of a scatterplot.

`geom_smooth()` uses a model to fit a smoothed line to the data and then visualizes the results. By default, `geom_smooth()` fits a loess smooth to data sets with less than 1,000 observations, and a generalized additive model to data sets with more than 1,000 observations. 

```{r echo = FALSE, out.width="50%", message = FALSE, warning = FALSE}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot() +
    geom_point(mapping = aes(x = mean_cty, y = mean_hwy)) +
    ggtitle("geom_point)") +
    ylim(16, 30)

mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot() +
    geom_smooth(mapping = aes(x = mean_cty, y = mean_hwy), se = FALSE) +
    ggtitle("geom_smooth()") +
    ylim(16, 30)
```

### method

You can use the `method` parameter of `geom_smooth()` to fit and display other types of model lines. To do this, pass `method` the name of an R modeling function for `geom_smooth()` to use, such as `lm` (for linear models) or `glm` (for generalized linear models).

In the code below, use `geom_smooth()` to draw the linear model line that fits the data.

```{r r3, exercise = TRUE}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot() 
```

```{r r3-solution}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot() +
    geom_smooth(mapping = aes(x = mean_cty, y = mean_hwy), method = lm)
```

```{r r3-check}
"Good job! Now let's look at a way to make geom_smooth() much more useful."
```

## Layers

### Add a layer

`geom_smooth()` becomes much more useful when you combine it with `geom_point()` to create a scatterplot that contains both:

* raw data
* a trend line

In ggplot2, you can add multiple geoms to a plot by adding multiple geom functions to the plot call. For example, the code below creates a plot that contains both points and a smooth line. Imagine what the results will look like in your head, and then run the code to see if you are right.

```{r r4, exercise = TRUE}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot() +
    geom_point(mapping = aes(x = mean_cty, y = mean_hwy)) +
    geom_smooth(mapping = aes(x = mean_cty, y = mean_hwy), method = lm) 
```

```{r r4-check}
"Good job! You can add as many geom functions as you like to a plot; but, in practice, a plot will become hard to interpret if it contains more than two or three geoms."
```

### geom_label_repel()

Do you remember how the labels that we made early overlapped each other and ran off our graph? The `geom_label_repel()` geom from the ggrepel package mitigates these problems by using an algorithm to arrange labels within a plot. It works best in conjunction with a layer of points that displays the true location of each observation.

Use `geom_label_repel()` to add a new layer to our plot below. `geom_label_repel()` requires the same aesthetics as `geom_label()`: x, y, and label (here set to class).

```{r r5, exercise = TRUE}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot() +
    geom_point(mapping = aes(x = mean_cty, y = mean_hwy)) +
    geom_smooth(mapping = aes(x = mean_cty, y = mean_hwy), method = lm)
```

```{r r5-solution}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot() +
    geom_point(mapping = aes(x = mean_cty, y = mean_hwy)) +
    geom_smooth(mapping = aes(x = mean_cty, y = mean_hwy), method = lm) +
    geom_label_repel(mapping = aes(x = mean_cty, y = mean_hwy, label = class))
```

```{r r5-check}
"Good job! The ggrepel package also provides geom_text_repel(), which is an analog for geom_text()."
```

### Code duplication

If you study the solution for the previous exercise, you'll notice a fair amount of duplication. We set the same aesthetic mappings in three different places.

```{r eval = FALSE}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot() +
    geom_point(mapping = aes(x = mean_cty, y = mean_hwy)) +
    geom_smooth(mapping = aes(x = mean_cty, y = mean_hwy), method = lm) +
    geom_label_repel(mapping = aes(x = mean_cty, y = mean_hwy, label = class))
```

You should try to avoid duplication whenever you can in code because duplicated code invites typos, is hard to update, and takes longer than needed to write. Thankfully, ggplot2 provides a way to avoid duplication across multiple layers.

### ggplot() mappings

You can set aesthetic mappings in two places within any ggplot2 call. You can set the mappings inside of a geom function, as we've been doing. Or you can set the mappings inside of the `ggplot()` function like below:

```{r out.width = "80%", message = FALSE, warning = FALSE, message = FALSE, warning = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point()
```

### Global vs. Local mappings

ggplot2 will treat any mappings set in the `ggplot()` function as _global_ mappings. Each layer in the plot will inherit and use these mappings. 

ggplot2 will treat any mappings set in a geom function as _local_ mappings. Only the local layer will use these mappings. The mappings will override the global mappings if the two conflict, or add to them if they do not.

This system creates an efficient way to write plot calls:

```{r out.width = "80%", message = FALSE, warning = FALSE, warning = FALSE, message = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(mapping = aes(color = class), se = FALSE)
```

### Exercise 2

Reduce duplication in the code below by moving as many local mappings into the global mappings as possible. Rerun the new code to ensure that it creates the same plot.

```{r r6, exercise = TRUE, exercise.eval = TRUE}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot() +
    geom_point(mapping = aes(x = mean_cty, y = mean_hwy)) +
    geom_smooth(mapping = aes(x = mean_cty, y = mean_hwy), method = lm) +
    geom_label_repel(mapping = aes(x = mean_cty, y = mean_hwy, label = class))
```

```{r r6-solution}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy)) %>% 
  ggplot(mapping = aes(x = mean_cty, y = mean_hwy)) +
    geom_point() +
    geom_smooth(method = lm) +
    geom_label_repel(mapping = aes(label = class))
```

```{r r6-check}
"Good job! Remember that not every mapping should be a global mapping. Here, only geom_label_repel uses the label aesthetic. Hence, it should remain a local aesthetic to avoid unintended side effects, warnings, or errors."
```


### Exercise 3 - Global vs. Local

Recreate the plot below in the most efficient way possible.

```{r echo = FALSE, out.width = "80%", message = FALSE, warning = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = class)) +
  geom_smooth()
```

```{r r7, exercise = TRUE}

```

```{r r7-solution}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = class)) +
  geom_smooth()
```

```{r r7-check}
"Good Job!"
```

### Global vs. Local data

The `data` argument also follows a global vs. local system. If you set the `data` argument of a geom function, the geom will use the data you supply instead of the data contained in `ggplot()`. This is a convenient way to highlight groups of points.

Use `data` arguments to recreate the plot below. I've started the code for you.

```{r echo = FALSE, out.width = "80%", message = FALSE, warning = FALSE}
mpg2 <- filter(mpg, class == "2seater")

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_point(data = mpg2, color = "red", size = 2)
```

```{r r8, exercise = TRUE, }
mpg2 <- filter(mpg, class == "2seater")
```

```{r r8-solution}
mpg2 <- filter(mpg, class == "2seater")

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_point(data = mpg2, color = "red", size = 2)
```

```{r r8-check}
"Good Job!"
```

### Exercise 4 - Global vs. Local data

Use `data` arguments to recreate the plot below.

```{r echo = FALSE, out.width = "80%", message = FALSE, warning = FALSE}
mpg3 <- filter(mpg, hwy > 40)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_label_repel(data = mpg3, mapping = aes(label = class))
```

```{r r9, exercise = TRUE}

```

```{r r9-solution}
mpg3 <- filter(mpg, hwy > 40)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_label_repel(data = mpg3, mapping = aes(label = class))
```

```{r r9-check}
"Good Job!"
```

### last_plot()

When exploring data, you'll often make a plot and then think of a way to improve it. Instead of starting from scratch or copying and pasting your code, you can use ggplot2's `last_plot()` function. `last_plot()` returns the most recent plot call, which makes it easy to build up a plot one layer at a time.

```{r out.width = "80%", message = FALSE, warning = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point()

last_plot() +
  geom_smooth()

last_plot() +
  geom_smooth(method = lm, color = "purple")
```

### Saving plots

If you'd like to work with a plot later, you can save it to an R object. Later you can display the plot or add to it, as if you were using `last_plot()`. 

```{r}
p <- ggplot(data = mpg) +
  geom_point(mapping = aes(x = cty, y = hwy))
```

Notice that ggplot2 will not display a plot when you save it. It waits until you call the saved object.

```{r out.width = "80%", message = FALSE, warning = FALSE}
p
```


### geom_rug()

`geom_rug()` adds another type of summary to a plot. It uses displays the one dimensional marginal distributions of each variable in the scatterplot. These appear as collections of tickmarks along the $x$ and $y$ axes.

In the chunk below, use the `faithful` dataset to create a scatterplot that has the `waiting` variable on the $x$ axis and the `eruptions` variable on the $y$ axis. Use `geom_rug()` to add a rug plot to the scatterplot. Like `geom_point()`, `geom_rug()` requires x and y aesthetic mappings.

```{r r10, exercise = TRUE}

```

```{r r10-solution}
ggplot(data = faithful, mapping = aes(x = waiting, y = eruptions)) +
  geom_point() +
  geom_rug()
```

```{r r10-check}
'Good job! Pass geom_rug() the parameter sides = "l" to limit the rug plot to just the left-hand axis (y) or sides = "b" to limit the rug plot to just the bottom axis (x).'
```


### geom_jitter()

`geom_jitter()` plots a scatterplot and then adds a small amount of random noise to each point in the plot. It is a shortcut for adding a "jitter" position adjustment to a points plot (i.e, `geom_point(position = "jitter")`).

Why would you use `geom_jitter()`? Jittering provides a simple way to inspect patterns that occur in heavily gridded or overlapping data. To see what I mean, replace `geom_point()` with `geom_jitter()` in the plot below.

```{r r11, exercise = TRUE, exercise.eval = TRUE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = class, y = hwy))
```

```{r r11-solution}
ggplot(data = mpg) +
  geom_jitter(mapping = aes(x = class, y = hwy))
```

```{r r11-check}
"Good job! You can also jitter in only a single direction. To turn off jittering in the x direction set width = 0 in geom_jitter(). To turn off jittering in the y direction, set height = 0." 
```


### jitter and boxplots

`geom_jitter()` provides a convenient way to overlay raw data on boxplots, which display summary information.

Use the chunk below to create a boxplot of the previous graph. Arrange for the outliers to have an alpha of 0, which will make them completely transparent. Then add a layer of points that are jittered in $y$ direction, but not the $x$ direction.

```{r r12, exercise = TRUE}

```

```{r r12-solution}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_jitter(width = 0)
```

```{r r12-check}
"Good job! If you like, you can make the boxplots more visible by setting the alpha parameter of geom_jitter() to a low number, e.g. geom_jitter(mapping = aes(x = class, y = hwy), width = 0, alpha = 0.5)"
```


## Coordinate Systems

### coord_flip()

One way to customize a scatterplot is to plot it in a new coordinate system. ggplot2 provides several helper functions that change the coordinate system of a plot. You've already seen one of these in action in the [boxplots tutorial](https://tutorials.shinyapps.io/04-boxplots): `coord_flip()` flips the $x$ and $y$ axes of a plot.

```{r out.width = "80%", message = FALSE, warning = FALSE}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_jitter(width = 0) +
  coord_flip()
```

### The coord functions

Altogether, ggplot2 comes with seven coord functions:

* `coord_cartesian()` - (the default) Cartesian coordinates
* `coord_fixed()` - Cartesian coordinates that maintain a fixed aspect ratio as the plot window is resized
* `coord_flip()` - Cartesian coordinates with x and y axes flipped
* `coord_map()` and `coord_quickmap()` - cartographic projections for plotting maps
* `coord_polar()` - polar coordinates
* `coord_trans()` - transformed Cartesian coordinates

By default, ggplot2 will draw a plot in Cartesian coordinates unless you add one of the functions above to the plot code.

### coord_polar()

You use each coord function like you use `coord_flip()`, by adding it to a ggplot2 call.

So for example, you could add `coord_polar()` to a plot to make a graph that uses polar coordinates.

```{r out.width = "80%", message = FALSE, warning = FALSE}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = cut), width = 1) 

last_plot() +
  coord_polar()
```

### Coordinate systems and scatterplots

How can a coordinate system improve a scatterplot? 

Consider, the scatterplot below. It shows a strong relationship between the carat size of a diamond and its price. 

```{r echo = FALSE, out.width = "80%", message = FALSE, warning = FALSE}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price)) 
```

However, the relationship does not appear linear. It appears to have the form $y = x^{n}$, a common relationship found in nature. You can estimate the $n$ by replotting the data in a _log-log plot_.

### log-log plots

Log-log plots graph the log of $x$ vs. the log of $y$, which has a valuable visual effect. If you log both sides of a relationship like

$$y = x^{n}$$

You get a linear relationship with slope $n$:

$$log(y) = log(x^{n})$$
$$log(y) = n \cdot log(x)$$

In other words, log-log plots unbend power relationships into straight lines. Moreover, they display $n$ as the slope of the straight line, which is reasonably easy to estimate.

Try this by using the diamonds dataset to plot `log(carat)` against `log(price)`.

```{r r13, exercise = TRUE}

```
```{r r13-solution}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = log(carat), y = log(price))) 
```

```{r r13-check}
"Good job! Now let's look at how you can do the same transformation, and others as well with a coord function."
```

### coord_trans()

`coord_trans()` provides a second way to do the same transformation, or similar transformations.

To use `coord_trans()` give it an $x$ and/or a $y$ argument. Set each to the name of an R function surrounded by quotation marks. `coord_trans()` will use the function to transform the specified axis before plotting the raw data. 

```{r r14, exercise = TRUE}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price)) +
  coord_trans(x = "log", y = "log")
```


### Recap

Scatterplots are one of the most useful types of plots for data science. You will have many chances to use `geom_point()`, `geom_smooth()`, and `geom_label_repel()` in your day to day work.

However, this tutor introduced important two concepts that apply to more than just scatterplots:

* You can add **multiple layers** to any plot that you make with ggplot2
* You can add a different **coordinate system** to any plot that you make with ggplot2

<a class="typeform-share button" href="https://rstudio.typeform.com/to/zhIdjy" data-mode="popup" style="display:inline-block;text-decoration:none;background-color:#6699cc;color:white;cursor:pointer;font-family:Helvetica,Arial,sans-serif;font-size:14px;line-height:35px;text-align:center;margin:0;height:35px;padding:0px 23px;border-radius:3px;max-width:100%;white-space:nowrap;overflow:hidden;text-overflow:ellipsis;font-weight:normal;-webkit-font-smoothing:antialiased;-moz-osx-font-smoothing:grayscale;" target="_blank">Give Feedback </a> <script> (function() { var qs,js,q,s,d=document, gi=d.getElementById, ce=d.createElement, gt=d.getElementsByTagName, id="typef_orm_share", b="https://embed.typeform.com/"; if(!gi.call(d,id)){ js=ce.call(d,"script"); js.id=id; js.src=b+"embed.js"; q=gt.call(d,"script")[0]; q.parentNode.insertBefore(js,q) } })() </script>
