---
title: "Introduction to Iteration"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
---

```{r setup, include=FALSE, message = FALSE, warning=FALSE}
library(learnr)
library(tidyverse)

checker <- function(label, user_code, check_code, envir_result, evaluate_result, ...) {
  list(message = check_code, correct = TRUE, location = "append")
}
tutorial_options(exercise.timelimit = 60, exercise.checker = checker)
knitr::opts_chunk$set(error = TRUE, out.width = "100%")
```


## Iteration

### \t

Here is a number. You can do many things with it, like this:  

* Take the log of the number. Then click Submit Answer.

```{r number, exercise = TRUE}
1
```

```{r number-solution}
log(1)
```

```{r number-hint}
"You can take the log with `log()`"
```

```{r number-check}
"Good job. Don't worry, things will soon get harder."
```

### \t

Here are many numbers. What if you want to do things with them? Go ahead: 

* Take the log of each. Don't get fancy.

```{r numbers, exercise = TRUE}
1
2
3
4
5
```

```{r numbers-solution}
log(1)
log(2)
log(3)
log(4)
log(5)
```

```{r numbers-hint}
"You can call `log()` on each."
```

```{r numbers-check}
"Nice! You iterated through the numbers, calling log() on them one at a time.

But shouldn't your computer do this repetition for you? Why are you using a computer anyways?"
```

### Vectorization

You usually do not need to repeat yourself when you use R. Almost every function in R is _vectorized_. This means that if you put your values into a vector, and then run a function on the vector, R will apply the function iteratively to each of the values.

* Click Submit Answer to try it. It's worth seeing at least once!

```{r vectors, exercise = TRUE}
numbers <- c(1, 2, 3, 4, 5)
log(numbers)
```

```{r vectors-hint}
"Just click Submit Answer and see what happens."
```

```{r vectors-check}
"Vectorization in R works, but there is something that you should know. Click Continue to find out what it is."
```

### \t

Not every function in R is vectorized, and those that are vectorized often only work with _non-recursive vectors_.

What does that mean? Well, there are two types of vectors in R:

1. non-recursive vectors
1. recursive vectors

Let's look at each.

### Non-recursive vectors

Non-recursive vectors are usually what you think about when you think of a vector. Each element of a non-recursive vector is a single value, and all of the values in a non-recursive vector are the same type. Examples of non-recursive vectors in R include:

* double vectors (i.e. numeric vectors)
* integer vectors
* character vectors
* logical vectors
* factors
* and so on

### \t

Almost every function in R is vectorized to work with non-recursive vectors, as you saw with `log()`. Here, `numbers` is a non-recursive vector.

```{r echo = TRUE}
numbers <- c(1, 2, 3, 4, 5)
log(numbers)
```

What about recursive vectors?

### Recursive vectors

A _recursive vector_ is a vector whose individual elements can be anything, even other vectors, even other recursive vectors. Hence, the name.

Recursive vectors are so special that R gives them a special name, _lists_. You can think of lists as the glue that holds R together. R builds data frames, environments, model objects and more by grouping together different types of non-recursive vectors with a list.

### \t

But are R functions vectorized to work with lists? 

* Try running `log()` on the list below to find out. Click Run Code.

```{r lists, exercise = TRUE}
list1 <- list(1, 2, 3, 4, 5)

log(list1)
```

```{r lists-hint}
"Just click Submit Answer and see what happens."
```

### \t

Almost no function in R is vectorized to work with lists. And you can see why: lists can contain different types of objects, as `list2` demonstrates. The second element of `list2` is a character vector. 

```{r}
chars <- c("a", "b", "c")
list2 <- list(1, chars)
list2
```

How likely is it that an R function will work with both `1` and the vector `c("a", "b", "c")`? 

Non-recursive vectors always contain a single type of data, but lists often contain many types of data. Most R functions only work with a single type of data, so their authors arrange for them to handle non-recursive vectors, but not lists.

### Important vocabulary

Recursive vectors have an everyday name, lists, but non-recursive vectors do not. People tend to call non-recursive _vectors_, but that is not very accurate since recursive vectors are vectors too.

To distinguish non-recursive vectors from recursive vectors, I will refer to recursive vectors as _lists_ and non-recursive vectors as _simple vectors_.

It's important to keep in mind that lists _are_ a type of vector. So if you see me say _vector_ instead of _simple vector_, I mean both lists and simple vectors.

Now, how should we work with lists?

### What should you do?

Don't worry, there's no reason to go back to repeating things by hand for lists, although you could. Couldn't you? Let's see. 

You can extract each element of a list with the `[[i]]` bracket syntax. For example, `list2[[2]]` extracts the second element of `list2`:

```{r}
list2[[2]]
```

* Finish the code below to take the log of each value in `list1`. Then click Submit Answer.

```{r manual-setup}
list1 <- list(1, 2, 3, 4, 5)
```

```{r manual, exercise = TRUE}
list1[[1]]
list1[[2]]
list1[[3]]
list1[[4]]
list1[[5]]
```

```{r manual-hint-1}
"Apply `log()` to each list component below."
```

```{r manual-hint-2}
"You can pass each component straight into the `log()` function, e.g. `log(list1[[1]])`."
```

```{r manual-solution}
log(list1[[1]])
log(list1[[2]])
log(list1[[3]])
log(list1[[4]])
log(list1[[5]])
```

```{r manual-check}
"Nice! But not nice enough to use in daily practice. Let's learn how to tell your computer to iterate over lists for you."
```

### What about for loops?

If you've written a for loop before, you might be saying to yourself, "Isn't map() doing the same thing as a for loop?" (If you've never written for loops before, you can skip ahead).

Yes the map functions _are_ doing the same thing as a for loop, and bravo for making that connection!

So why use `map()` instead of `for()`?

### \t

You certainly can use `for()`, but the map functions provide some benefits:

1. They implement the for loop for you, saving you time and brain power
1. They work inside dplyr functions like `mutate()` and `summarise()` (Have you ever tried to implement a for loop inside of `mutate()`? It becomes important when you work with list columns.)

### But wait...

Okay, that sounds good. But if you've been using R for awhile you might have a follow up question: "What about `lapply()`, `sapply()` and `vapply()`? Don't they do the same thing as `map()`?" 

Yes they do; and again, good job making the connection! Let's wait until the More Map tutorial to answer this question.