---
title: "Data Visualization Basics"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
tutorial:
  id: "com.rstudio.primers.02-vis-basics"
  version: 1.0
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
checker <- function(label, user_code, check_code, envir_result, evaluate_result, ...) {
  list(message = check_code, correct = TRUE, location = "append")
}
tutorial_options(exercise.timelimit = 60, exercise.checker = checker)
```

```{r, context="server-start", include=FALSE}
# Capture metrics only if running an official primer hosted by RStudio
library(curl)
library(later)
tryCatch(
  source("https://metrics.rstudioprimers.com/learnr/installMetrics", local=TRUE), 
  error = function(e){ 
    print("Warning: An error occurred with the tracking code.")
  }
)
```

```{r, context = "render", results = 'asis', echo = FALSE}
tryCatch(
  source("https://metrics.rstudioprimers.com/learnr/installClient", local=TRUE)$value,
  error = function(e){ 
    print("Warning: An error occurred with the client code.")
  }
)
```

## Welcome

### 

Visualization is one of the most important tools for data science. 

It is also a great way to start learning R; when you visualize data, you get an immediate payoff that will keep you motivated as you learn. Afterall, learning a new language can be hard!

This tutorial will teach you how to visualize data with R's most popular visualization package, `ggplot2`.

### 

The tutorial focuses on three basic skills:

1. How to create graphs with a reusable **template**
1. How to add variables to a graph with **aesthetics**
1. How to make different "types" of graphs with **geoms** 

In this tutorial, we will use the [core tidyverse packages](http://tidyverse.org/), including `ggplot2`. I've already loaded the packages for you, so let's begin! 

***

These examples are excerpted from _R for Data Science_ by Hadley Wickham and Garrett Grolemund, published by O’Reilly Media, Inc., 2016, ISBN: 9781491910399. You can purchase the book at [shop.oreilly.com](http://shop.oreilly.com/product/0636920034407.do).


## A code template

> "The simple graph has brought more information to the data analyst’s mind 
> than any other device. " --- John Tukey

### 

Let's begin with a question to explore.

```{r predict, echo = FALSE}
question("What do you think: Do cars with bigger engines use more fuel than cars with smaller engines?", 
         type = "single",
         correct = "Great!",
         answer("Cars with bigger engines use _more_ fuel.", correct = TRUE, message = "In other words, there is a positive relationship between engine size and fuel efficiency. Now let's test your hypothesis against data."),
         answer("Cars with bigger engines use _less_ fuel.", correct = TRUE, message = "In other words, there is a negative relationship between engine size and fuel efficiency. Now let's test your hypothesis with data.")
)
```

### `mpg`

You can test your hypothesis with the `mpg` dataset that comes in the `ggplot2` package. `mpg` contains observations collected on 38 models of cars by the US Environmental Protection Agency. 

To see the `mpg` data frame, type `mpg` in the code block below and click "Submit Answer".

```{r mpg0, exercise = TRUE}

```

```{r mpg0-solution}
mpg
```

```{r mpg0-check}
"Good job!  We'll use interactive code chunks like this throughout these tutorials. Whenever you encounter one, you can click Submit Answer to run (or re-run) the code in the chunk. If there is a Solution button, you can click it to see the answer."
```

You can use the black triangle that appears at the top right of the table to scroll through all of the columns in `mpg`.

### 

Among the variables in `mpg` are:

1. `displ`, a car's engine size, in liters.
1. `hwy`, a car's fuel efficiency on the highway, in miles per gallon (mpg). A car with a low mpg consumes more fuel than a car with a high mpg when they travel the same distance.

Now let's use this data to make our first graph.

### A plot

The code below uses functions from the **ggplot2** package to plot the relationship between `displ` and `hwy`. 

To see the plot, click "Run Code."

```{r ggplot1, exercise = TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

Can you spot the relationship?

### And the answer is...

The plot shows a negative relationship between engine size (`displ`) and fuel efficiency (`hwy`). Points that have a large value of `displ` have a small value of `hwy` and vice versa.

In other words, cars with big engines use _more_ fuel. If that was your hypothesis, you were right!

Now let's look at how we made the plot.

### `ggplot()`

Here's the code that we used to make the plot. Notice that it contains three functions: `ggplot()`, `geom_point()`, and `aes()`.

```{r eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

In R, a function is a name followed by a set of parentheses. Many functions require special information to do their jobs, and you write this information between the parentheses.

### `ggplot`

The first function, `ggplot()`, creates a coordinate system that you can add layers to. The first argument of `ggplot()` is the dataset to use in the graph. 

By itself, `ggplot(data = mpg)` creates an empty graph, which looks like this. 

```{r}
ggplot(data = mpg)
```

### geom_point()

`geom_point()` adds a layer of points to the empty plot created by `ggplot()`. This gives us a scatterplot. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

### mapping = aes()

`geom_point()` takes a `mapping` argument that defines which variables in your dataset are mapped to which axes in your graph. The `mapping` argument is always paired with the function `aes()`, which you use to gather together all of the mappings that you want to create. 

### 

Here, we want to map the `displ` variable to the x axis and the `hwy` variable to the y axis, so we add `x = displ` and `y = hwy` inside of `aes()` (and we separate them with a comma). 

Where will ggplot2 look for these mapped variables? In the data frame that we passed to the `data` argument, in this case, `mpg`.

### A graphing workflow

Our code follows the common workflow for making graphs with ggplot2. To make a graph, you: 

1. Start the graph with `ggplot()`
1. Add elements to the graph with a `geom_` function
1. Select variables with the `mapping = aes()` argument

### A graphing template

In fact, you can turn our code into a reusable template for making graphs. To make a graph, replace the bracketed sections in the code below with a data set, a `geom_` function, or a collection of mappings.

Give it a try! Replace the bracketed sections with `mpg`, `geom_boxplot`, and `x = class, y = hwy` to make a slightly different graph. Be sure to delete the `#` symbols before you run the code.

```{r template, exercise = TRUE, eval = FALSE}
# ggplot(data = <DATA>) + 
#  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

```{r template-solution}
ggplot(data = mpg) + 
  geom_boxplot(mapping = aes(x = class, y = hwy))
```

```{r template-check}
"Good job! This plot uses boxplots to compare the fuel efficiencies of different types of cars. ggplot2 comes with many geom functions that each add a different type of layer to a plot. You'll learn more about boxplots and other geoms in the tutorials that follow."
```

### Common problems

As you start to run R code, you're likely to run into problems. Don't worry --- it happens to everyone. I have been writing R code for years, and every day I still write code that doesn't work! 

Start by carefully comparing the code that you're running to the code in the examples. R is extremely picky, and a misplaced character can make all the difference. Make sure that every `(` is matched with a `)` and every `"` is paired with another `"`. Also pay attention to capitalization; R is case sensitive.

### + location

One common problem when creating ggplot2 graphics is to put the `+` in the wrong place: it has to come at the end of a line, not the start. In other words, make sure you haven't accidentally written code like this:

```R
ggplot(data = mpg) 
+ geom_point(mapping = aes(x = displ, y = hwy))
```

### help

If you're still stuck, try the help. You can get help about any R function by running `?function_name` in a code chunk, e.g. `?geom_point`. Don't worry if the help doesn't seem that helpful --- instead skip down to the bottom of the help page and look for a code example that matches what you're trying to do.

If that doesn't help, carefully read the error message that appears when you run your (non-working) code. Sometimes the answer will be buried there! But when you're new to R, you might not yet know how to understand the error message. Another great tool is Google: try googling the error message, as it's likely someone else has had the same problem, and has gotten help online.

### Exercise 1

Run `ggplot(data = mpg)` what do you see?

```{r blank, exercise = TRUE}

```

```{r blank-check}
"Good job! A ggplot that has no layers looks blank. To finish the graph, add a geom function."
```
     
### Exercise 2 

Make a scatterplot of `cty` vs `hwy`.

```{r scatterplot, exercise = TRUE}

```

```{r scatterplot-check}
"Excellent work!"
```

```{r scatterplot-hint}
"Scatterplots are also called points plots and bubble plots. They use the point geom."
```

### Exercise 3  

What happens if you make a scatterplot of `class` vs `drv`. Try it. Why is the plot not useful?
    
```{r scatterplot2, exercise = TRUE}

```

```{r scatterplot2-check}
"Nice job! `class` and `drv` are both categorical variables. As a result, points can only appear at certain values, where many points overlap each other. You have no idea how many points fall on top of each other at each location. Experiment with geom_count() to find a better solution."
```


## Aesthetic mappings

> "The greatest value of a picture is when it forces us to notice what we
> never expected to see." --- John Tukey


### A closer look

In the plot below, one group of points (highlighted in red) seems to fall outside of the linear trend between engine size and gas mileage. These cars have a higher mileage than you might expect. How can you explain these cars? 

```{r, echo = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() + 
  geom_point(data = dplyr::filter(mpg, displ > 5, hwy > 20), colour = "red", size = 2.2)
```

### A hypothesis

Let's hypothesize that the cars are hybrids. One way to test this hypothesis is to look at the `class` value for each car. The `class` variable of the `mpg` dataset classifies cars into groups such as compact, midsize, and SUV. If the outlying points are hybrids, they should be classified as compact cars or, perhaps, subcompact cars (keep in mind that this data was collected before hybrid trucks and SUVs became popular). To check this, we need to add the `class` variable to the plot.

### Aesthetics

You can add a third variable, like `class`, to a two dimensional scatterplot by mapping it to a new __aesthetic__. An aesthetic is a visual property of the objects in your plot. Aesthetics include things like the size, the shape, or the color of your points. 

You can display a point (like the one below) in different ways by changing the values of its aesthetic properties. Since we already use the word "value" to describe data, let's use the word "level" to describe aesthetic properties. Here we change the levels of a point's size, shape, and color to make the point small, triangular, or blue:

```{r, echo = FALSE, asp = 1/4, out.width = "100%"}
ggplot() +
  geom_point(aes(1, 1), size = 20) +
  geom_point(aes(2, 1), size = 10) + 
  geom_point(aes(3, 1), size = 20, shape = 17) + 
  geom_point(aes(4, 1), size = 20, colour = "blue") + 
  scale_x_continuous(NULL, limits = c(0.5, 4.5), labels = NULL) + 
  scale_y_continuous(NULL, limits = c(0.9, 1.1), labels = NULL) + 
  theme(aspect.ratio = 1/3)
```

### A strategy

We can add the `class` variable to the plot by mapping the levels of an aesthetic (like color) to the values of `class`. For example, we can color a point green if it belongs to the compact class, blue if it belongs to the midsize class, and so on. 

Let's give this a try. Fill in the blank piece of code below with `color = class`. What happens? Delete the commenting symbols (`#`) before running your code. (If you prefer British English, you can use `colour` instead of `color`.)

```{r aesthetics, exercise = TRUE, exercise.eval = FALSE}
# ggplot(data = mpg) + 
#   geom_point(mapping = aes(x = displ, y = hwy, ____________))
```

<div id="aesthetics-hint">
**Hint:** Be sure to remove all of the underlines and hashtags from the code.
</div>

```{r aesthetics-solution}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

```{r aesthetics-check}
"Great Job! You can now tell which class of car each point represents by examining the color of the point."
```

### And the answer is...

The colors reveal that many of the unusual points in `mpg` are two-seater cars. These cars don't seem like hybrids, and are, in fact, sports cars! Sports cars have large engines like SUVs and pickup trucks, but small bodies like midsize and compact cars, which improves their gas mileage. In hindsight, these cars were unlikely to be hybrids since they have large engines.

This isn't the only insight we've gleaned; you've also learned how to add new aesthetics to your graph. Let's review the process.

### Aesthetic mappings

To map an aesthetic to a variable, set the name of the aesthetic equal to the name of the variable, and do this inside `mapping = aes()`. ggplot2 will automatically assign a unique level of the aesthetic (here a unique color) to each unique value of the variable. ggplot2 will also add a legend that explains which levels correspond to which values.

This insight gives us a new way to think about the mapping argument. Mappings tell ggplot2 more than which variables to put on which axes, they tell ggplot2 which variables to map to which visual properties. The x and y locations of each point are just two of the many visual properties displayed by a point.

### Other aesthetics

In the above example, we mapped color to `class`, but we could have mapped size to `class` in the same way. 

Change the code below to map `size` to `class`. What happens? 

```{r size, exercise = TRUE, exercise.eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

<div id="size-hint">
**Hint:** If `color` controls the _aesthetic_, what word do you suppose controls the _size_ aesthetic?
</div>

```{r size-solution}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

```{r size-check}
"Great Job! Now the size of a point represents its class. Did you notice the warning message? ggplot2 gives us a warning here because mapping an unordered variable (class) to an ordered aesthetic (size) is not a good idea."
```

### alpha

You can also map `class` to the `alpha` aesthetic, which controls the transparency of the points. Try it below.

```{r alpha, exercise = TRUE, exercise.eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

<div id="alpha-hint">
**Hint:** If `color` controls the _aesthetic_, what word do you suppose controls the _alpha_ aesthetic?
</div>

```{r alpha-solution}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

```{r alpha-check}
"Great Job! If you look closely, you can spot something subtle: many locations contain multiple points stacked on top of each other (alpha is additive so multiple transparent points will appear opaque)."
```

### Shape

Let's try one more aesthetic. This time map the class of the points to `shape`, then look for the SUVs. What happened?

```{r shape, exercise = TRUE, exercise.eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

<div id="shape-hint">
**Hint:** If `color` controls the _aesthetic_, what word do you suppose controls the _shape_ aesthetic?
</div>

```{r shape-solution}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

```{r shape-check}
"Good work! What happened to the SUVs? ggplot2 will only use six shapes at a time. By default, additional groups will go unplotted when you use the shape aesthetic. So only use it when you have fewer than seven groups."
```

### Exercise 1

In the code below, map `cty`, which is a continuous variable, to `color`, `size`, and `shape`. How do these aesthetics behave differently for continuous variables, like `cty`, vs. categorical variables, like `class`? 

```{r cty, exercise = TRUE, exercise.eval = FALSE}
# Map cty to color
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

# Map cty to size
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

# Map cty to shape
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

```{r cty-check}
"Very nice! ggplot2 treats continuous and categorical variables differently. Noteably, ggplot2 supplies a blue gradient when you map a continuous variable to color, and ggplot2 will not map continuous variables to shape."
```
    
### Exercise 2

Map `class` to `color`, `size`, and `shape` all in the same plot. Does it work?

```{r multiple, exercise = TRUE, exercise.eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

```{r multiple-check}
"Very nice! ggplot2 can map the same variable to multiple aesthetics."
```
    
<div id="multiple-hint">
**Hint:** Be sure to set each mapping separately, e.g. `color = class`, `size = class`, etc.
</div>    
    
### Exercise 3

What happens if you map an aesthetic to something other than a variable name, like `aes(colour = displ < 5)`?  Try it.
    
```{r nonvar, exercise = TRUE, exercise.eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

```{r nonvar-check}
"Good job! ggplot2 will map the aesthetic to the results of the expression. Here, ggplot2 mapped the color of each point to TRUE or FALSE based on whther the point's `displ` value was less than five."
```

### Setting aesthetics

What if you just want to make all of the points in your plot blue, like in the plot below?

```{r echo = FALSE, out.width="80%"}
ggplot(mpg) +
  geom_point(aes(displ, hwy), color = "blue")
```

You can do this by setting the color aesthetic _outside_ of the `aes()` function, like this

```{r setting, exercise = TRUE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

### Setting vs. Mapping

Setting works for every aesthetic in ggplot2. If you want to manually set the aesthetic to a **value** in the visual space, set the aesthetic outside of `aes()`. 

```{r setting2, exercise = TRUE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue", shape = 3, alpha = 0.5)
```

If you want to map the aesthetic to a **variable** in the data space, map the aesthetic inside `aes()`.

```{r setting3, exercise = TRUE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class, shape = fl, alpha = displ))
```

### Exercise 4

What do you think went wrong in the code below? Fix the code so it does something sensible.

```{r e4, exercise = TRUE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

```{r e4-solution}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

```{r e4-check}
"Good job! Putting an aesthetic in the wrong location is one of the most common graphing errors. Sometimes it helps to think of legends. If you will need a legend to understand what the color/shape/etc. means, then you should probably put the aesthetic inside `aes()` --- ggplot2 will build a legend for every aesthetic mapped here. If the aesthetic has no meaning and is just... well, aesthetic, then set it outside of `aes()`."
```


### Recap

For each aesthetic, you associate the name of the aesthetic with a variable to display, and you do this within `aes()`. 

Once you map a variable to an aesthetic, ggplot2 takes care of the rest. It selects a reasonable scale to use with the aesthetic, and it constructs a legend that explains the mapping between levels and values. For x and y aesthetics, ggplot2 does not create a legend, but it creates an axis line with tick marks and a label. The axis line acts as a legend; it explains the mapping between locations and values.

You've experimented with the most common aesthetics for points: `x`, `y`, `color`, `size`, `alpha` and `shape`. Each geom uses its own set of aesthetics (you wouldn't expect a line to have a shape, for example). To find out which aesthetics a geom uses, open its help page, e.g. `?geom_line`.

This raises a new question that we've only brushed over: what is a geom?

## Geometric objects

### Geoms

How are these two plots similar? 

```{r echo = FALSE, out.width = "50%", fig.align="default", message = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

Both plots contain the same x variable, the same y variable, and both describe the same data. But the plots are not identical. Each plot uses a different visual object to represent the data. In ggplot2 syntax, we say that they use different __geoms__.

A __geom__ is the geometrical object that a plot uses to represent observations. People often describe plots by the type of geom that the plot uses. For example, bar charts use bar geoms, line charts use line geoms, boxplots use boxplot geoms, and so on. Scatterplots break the trend; they use the point geom. 

As we see above, you can use different geoms to plot the same data. The plot on the left uses the point geom, and the plot on the right uses the smooth geom, a smooth line fitted to the data. 

### Geom functions

To change the geom in your plot, change the geom function that you add to `ggplot()`. For example, take this code which makes the plot on the left (above), and change `geom_point()` to `geom_smooth()`. What do you get? 

```{r geom, exercise = TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

```{r geom-solution}
# right
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

```{r geom-check}
"Good job! You get the plot on the right (above)."
```

### More about geoms

ggplot2 provides over 30 geom functions that you can use to make plots, and extension packages provide even more (see <https://exts.ggplot2.tidyverse.org/gallery/> for a sampling). You'll learn how to use these geoms to explore data in the [Visualize Data]() primer. 

Until then, the best way to get a comprehensive overview of the available geoms is with the [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf). To learn more about any single geom, look at its help page, e.g. `?geom_smooth`.

### Exercise 1

What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

### Exercise 2  


```{r q1, echo = FALSE}
question("What does the `se` argument to `geom_smooth()` do?",
         answer("Nothing. `se` is not an argument of `geom_smooth()`"),
         answer("chooses a method for calculating the smooth line", message = "That is what the `method` argument of `geom_smooth()` does."),
         answer("controls whether or not to **s**how **e**rrors"),
         answer("Adds or removes a standard error ribbon around the smooth line", correct = TRUE),
         allow_retry = TRUE
)
```

### Putting it all together

The ideas that you've learned here: geoms, aesthetics, and the implied existence of a data space and a visual space combine to form a system known as the Grammar of Graphics.

The Grammar of Graphics provides a systematic way to build any graph, and it underlies the ggplot2 package. In fact, the first two letters of ggplot2 stand for "Grammar of Graphics".

### The Grammar of Graphics

The best way to understand the Grammar of Graphics is to see it explained in action:

![](https://vimeo.com/223812632)

## The ggplot2 package

### What is a package?

Throughout this tutorial, I've referred to ggplot2 as a package. What does that mean?

The R language is subdivided into __packages__, small collections of data sets and functions that all focus on a single task. The functions that we used in this tutorial come from one of those packages, the ggplot2 package, which focuses on visualizing data.

### What should you know about packages?

When you first install R, you get a small collection of core packages known as __base R__. The remaining packages---there are over 10,000 of them---are optional. You don't need to install them unless you want to use them.

ggplot2 is one of these optionals packages, so are the other packages that we will look at in these tutorials. Some of the most popular and most modern parts of R come in the optional packages.

You don't need to worry about installing packages in these tutorials. Each tutorial comes with all of the packages that you need pre-installed; this is how we make the tutorials easy to use.

However, one day, you may want to use R outside of these tutorials. When that day comes, you'll want to remember which packages to download to acquire the functions you use here. Throughout the tutorials, I will try to make it as clear as possible where each function comes from!

If you'd like to learn more about installing R packages (or R or the RStudio IDE), the [Set Up](http://tutorials.shinyapps.io/00-setup) video tutorial walks you through the process of setting up R on your own computer.

### Where to from here

Congratulations! You can use the ggplot2 code template to plot any dataset in many different ways. As you begin exploring data, you should incorporate these tools into your workflow.

There is much more to ggplot2 and Data Visualization than we have covered here. If you would like to learn more about visualizing data with ggplot2, check out RStudio's primer on [Data Visualization]().

Your new data visualization skills will make it easier to learn other parts of R, because you can now visualize the results of any change that you make to data. you'll put these skills to immediate use in the next tutorial, which will show you how to extract values from datasets, as well as how to compute new variables and summary statistics from your data. See you there.
